---
title: "meta_analysis"
author: "Killian"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Preparation
## 1.1 Setting seed for reproducibility

```{r}
set.seed(42)
```

## 1.2 Loading packages

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(ggplot2)
library(edgeR)
library(biomaRt)
library(metapro)
library(ggVennDiagram)
library(UpSetR)
library(DT)
library(clusterProfiler)
library(pathview)
library(gage)
library(gageData)
library(grid)
library(GO.db)
library(patchwork)
library(scales)
```

# 2. Preparation of the data
## 2.1 Loading count data and metadata

```{r}
# Read pre-processed counts and metadata
paired <- read.delim("counts_paired.txt", comment.char = "#", row.names = 1)
single <- read.delim("counts_single.txt", comment.char = "#", row.names = 1)
counts <- cbind(paired[, -(1:5)], single[rownames(paired), -(1:5)])
colnames(counts) <- sub(".*\\.([ES]RR\\d+)_.*", "\\1", colnames(counts))
metadata <- read.csv("metadata.csv", row.names = 1)

# Order counts to match metadata
counts <- counts[, metadata$Run]

# Split counts by study
campbell_counts <- counts[, metadata$Run[metadata$study == "campbell"], drop = FALSE]
huang_counts    <- counts[, metadata$Run[metadata$study == "huang"],    drop = FALSE]
smith_counts    <- counts[, metadata$Run[metadata$study == "smith"],    drop = FALSE]
morris_counts   <- counts[, metadata$Run[metadata$study == "morris"],   drop = FALSE]
```

## 2.2 Loading annotation

```{r}
ft <- read.delim("GCF_047663525.1_IASCAAS_PekinDuck_T2T_feature_table.txt",
                 quote = "", stringsAsFactors = FALSE)
annotation_data <- ft %>%
  filter(X..feature == "gene") %>%
  mutate(key = if_else(locus_tag != "" & !is.na(locus_tag), locus_tag, symbol)) %>%
  transmute(key, symbol, GeneID, class, name, chromosome, start, end, strand) %>%
  distinct() %>%
  group_by(key) %>%
  mutate(ensembl_gene_id = if_else(row_number() == 1, key,
                                   paste0(key, "_", row_number() - 1))) %>%
  ungroup() %>%
  transmute(ensembl_gene_id,
            external_gene_name = symbol,
            entrezgene_id      = GeneID,
            gene_biotype       = class,
            description        = name,
            chromosome, start, end, strand) %>%
  as.data.frame()

setdiff(rownames(counts), annotation_data$ensembl_gene_id)   # expect character(0)

annotation_data <- annotation_data[match(rownames(counts), annotation_data$ensembl_gene_id), ]
stopifnot(identical(rownames(counts), annotation_data$ensembl_gene_id))
```
## 2.3 Creating DGEList objects

```{r}
# Ensure metadata matches counts columns exactly
metadata <- metadata %>%
  filter(Run %in% colnames(counts)) %>%
  .[match(colnames(counts), .$Run), ] %>%
  mutate(sample_id = paste(dpi, treatment, biological_replicate, sep = "_"))

# Full DGEList for initial QC
y <- DGEList(counts = counts, samples = metadata)

# Per-study DGELists
campbell <- DGEList(counts = campbell_counts, samples = metadata %>% filter(study == "campbell"), genes = annotation_data)
huang    <- DGEList(counts = huang_counts,    samples = metadata %>% filter(study == "huang"),    genes = annotation_data)
morris   <- DGEList(counts = morris_counts,   samples = metadata %>% filter(study == "morris"),   genes = annotation_data)
smith    <- DGEList(counts = smith_counts,    samples = metadata %>% filter(study == "smith"),    genes = annotation_data)

rm(campbell_counts, huang_counts, morris_counts, smith_counts)
```

# 3. Visualisation, filtering, normalisation
## 3.1 Raw data PCA

```{r, message=FALSE, warning=FALSE}
# Log-CPM PCA on all samples
# Exclude Morris timepoints that are not part of the meta-analysis design
# (0.5 and 5 dpi only exist in the original Morris study and were never
# carried into the DE contrasts in section 4) so they neither appear in the
# plots below nor influence the computed principal components. This filtered
# copy is local to this chunk; the `morris` DGEList used downstream (3.2+)
# is untouched.
y_pca <- y[, !(y$samples$dpi %in% c(0.5, 5))]

pca_res <- prcomp(t(cpm(y_pca, log = TRUE)), scale. = TRUE)
pca <- merge(
  transform(as.data.frame(pca_res$x), Run = rownames(pca_res$x)),
  y_pca$samples,
  by = "Run"
)

# Add display labels
pca$study <- stringr::str_to_title(pca$study)
pca$treatment_label <- case_when(
  pca$study == "Huang"                         & pca$treatment == "HP"      ~ "Clade 2.3.4",
  pca$study %in% c("Smith", "Campbell")        & pca$treatment == "HP"      ~ "Clade 1",
  pca$study == "Morris"                        & pca$treatment == "HP_2.2"  ~ "Clade 2.2.2.1",
  pca$study == "Morris"                        & pca$treatment == "HP_2.3.2.1" ~ "Clade 2.3.2.1a",
  pca$treatment == "control"                                                 ~ "Control",
  TRUE                                                                       ~ pca$treatment
)

# PCA coloured by study, shaped by treatment
p_pca1 <- ggplot(pca, aes(PC1, PC2, color = study, shape = treatment_label)) +
  geom_point(size = 3) +
  stat_ellipse() +
  labs(x = "PC1", y = "PC2", shape = "Treatment", color = "Study") +
  theme_minimal()

print(p_pca1)
ggsave("supplementary_figure_1.png", p_pca1, width = 12, height = 8, dpi = 300)

# PCA coloured by DPI, faceted by treatment
ggplot(pca, aes(PC1, PC2, color = factor(dpi), shape = treatment_label)) +
  geom_point(size = 3) +
  stat_ellipse() +
  facet_wrap(~ treatment_label) +
  scale_color_viridis_d() +
  labs(x = "PC1", y = "PC2", shape = "Treatment", color = "DPI") +
  theme_minimal()

# Per-study PCA
for (s in unique(pca$study)) {
  pca_s <- na.omit(subset(pca, study == s))
  p <- ggplot(pca_s, aes(PC1, PC2, color = factor(dpi), shape = treatment_label)) +
    geom_point(size = 3) +
    scale_color_viridis_d() +
    labs(title = paste("PCA Plot - Study:", s), x = "PC1", y = "PC2") +
    theme_minimal()
  if (nrow(pca_s) >= 3) p <- p + stat_ellipse()
  print(p)
}

rm(pca_res, pca, p, pca_s, s, p_pca1, y_pca)
```

## 3.2 Sum technical replicates (Morris) and verify

```{r, message=FALSE}
cat("Morris dimensions before summing tech reps:", dim(morris), "\n")
morris <- sumTechReps(morris, ID = morris$samples$sample_id)
cat("Morris dimensions after summing tech reps:",  dim(morris), "\n")

# Quick PCA to verify
log_cpm <- cpm(morris, log = TRUE)
log_cpm <- log_cpm[apply(log_cpm, 1, var) > 0, ]
pca_morris <- data.frame(prcomp(t(log_cpm), scale. = TRUE)$x[, 1:2], morris$samples)

ggplot(pca_morris, aes(PC1, PC2, color = factor(dpi), shape = treatment)) +
  geom_point(size = 3) +
  stat_ellipse(level = 0.95) +
  scale_color_viridis_d() +
  labs(title = "PCA Plot of Morris Dataset (post tech-rep sum)", x = "PC1", y = "PC2") +
  theme_minimal()

rm(log_cpm, pca_morris)
```

## 3.3 Filtering low-count genes

```{r}
filter_and_report <- function(dge) {
  cat("Before filtering:", dim(dge), "\n")
  keep <- filterByExpr(dge, group = dge$samples$group)
  dge  <- dge[keep, , keep.lib.sizes = FALSE]
  cat("After filtering: ", dim(dge), "\n")
  dge
}

cat("campbell\n"); campbell <- filter_and_report(campbell)
cat("huang\n");    huang    <- filter_and_report(huang)
cat("morris\n");   morris   <- filter_and_report(morris)
cat("smith\n");    smith    <- filter_and_report(smith)

rm(filter_and_report)
```

## 3.4 Normalising

```{r}
campbell <- normLibSizes(campbell)
huang    <- normLibSizes(huang)
morris   <- normLibSizes(morris)
smith    <- normLibSizes(smith)
```

## 3.5 PCA after normalisation

```{r, message=FALSE, warning=FALSE}
plotPCA <- function(dge, title, exclude_dpi = NULL) {
  # exclude_dpi lets us drop timepoints that were never carried into the DE
  # contrasts (section 4) from the PCA plot only. Subsetting happens on a
  # local copy of `dge` inside this function, so the `campbell`/`huang`/
  # `morris`/`smith` DGELists used downstream for dispersion estimation and
  # differential expression are never modified.
  if (!is.null(exclude_dpi)) {
    dge <- dge[, !(dge$samples$dpi %in% exclude_dpi)]
  }
  log_cpm <- cpm(dge, log = TRUE)
  log_cpm <- log_cpm[apply(log_cpm, 1, var) > 0, ]
  pca <- data.frame(prcomp(t(log_cpm), scale. = TRUE)$x[, 1:2], dge$samples)
  ggplot(pca, aes(PC1, PC2, color = factor(dpi), shape = factor(treatment))) +
    geom_point(size = 3) +
    stat_ellipse(aes(group = interaction(dpi, treatment)), level = 0.95, linetype = "dashed") +
    ggtitle(title) +
    labs(color = "dpi", shape = "treatment") +
    theme_minimal()
}

# Print each per-study PCA into the knitted document, and additionally save
# a high-resolution PNG (300 dpi), same convention as the UpSet plots in
# section 5.6.
save_pca_plot <- function(p, basename, width_in = 8, height_in = 6) {
  print(p)
  ggsave(paste0(basename, ".png"), p, width = width_in, height = height_in, dpi = 300)
}

save_pca_plot(plotPCA(campbell, "Campbell: PCA Plot"), "supplementary_figure_2a_campbell_pca")
save_pca_plot(plotPCA(huang,    "Huang: PCA Plot"),    "supplementary_figure_2b_huang_pca")
save_pca_plot(plotPCA(morris,   "Morris: PCA Plot", exclude_dpi = c(0.5, 5)),
              "supplementary_figure_2c_morris_pca")
save_pca_plot(plotPCA(smith,    "Smith: PCA Plot"),    "supplementary_figure_2d_smith_pca")

rm(y, counts, plotPCA, save_pca_plot)
```

# 4. Differential expression analysis
## 4.1 Estimate dispersion and fit model

```{r}
make_design <- function(dge) {
  d <- model.matrix(~ 0 + group, data = dge$samples)
  colnames(d) <- gsub("group", "", colnames(d))
  d
}

design_campbell <- make_design(campbell)
campbell <- estimateDisp(campbell, design_campbell)
fit_campbell <- glmFit(campbell, design_campbell)

design_smith <- make_design(smith)
smith <- estimateDisp(smith, design_smith)
fit_smith <- glmFit(smith, design_smith)

design_morris <- make_design(morris)
morris <- estimateDisp(morris, design_morris)
fit_morris <- glmFit(morris, design_morris)

design_huang <- make_design(huang)
# Huang has only one condition so we borrow BCV from the other studies
average_dispersion <- mean(c(morris$common.dispersion, campbell$common.dispersion, smith$common.dispersion))
print(morris$common.dispersion)
print(campbell$common.dispersion)
print(smith$common.dispersion)
print(average_dispersion)
fit_huang <- glmFit(huang, design_huang, dispersion = average_dispersion)

rm(make_design, average_dispersion)
```

## 4.2 Differential expression

```{r}
contrasts_campbell <- makeContrasts(
  HPvsControl_1dpi = HP_1 - control_1,
  HPvsControl_2dpi = HP_2 - control_2,
  HPvsControl_3dpi = HP_3 - control_3,
  levels = design_campbell
)
campbell_1 <- topTags(glmLRT(fit_campbell, contrast = contrasts_campbell[, "HPvsControl_1dpi"]), n = Inf)$table
campbell_2 <- topTags(glmLRT(fit_campbell, contrast = contrasts_campbell[, "HPvsControl_2dpi"]), n = Inf)$table
campbell_3 <- topTags(glmLRT(fit_campbell, contrast = contrasts_campbell[, "HPvsControl_3dpi"]), n = Inf)$table

contrasts_huang <- makeContrasts(
  HPvsControl_1dpi = HP_1 - control_3,
  HPvsControl_2dpi = HP_2 - control_3,
  HPvsControl_3dpi = HP_3 - control_3,
  levels = design_huang
)
huang_1 <- topTags(glmLRT(fit_huang, contrast = contrasts_huang[, "HPvsControl_1dpi"]), n = Inf)$table
huang_2 <- topTags(glmLRT(fit_huang, contrast = contrasts_huang[, "HPvsControl_2dpi"]), n = Inf)$table
huang_3 <- topTags(glmLRT(fit_huang, contrast = contrasts_huang[, "HPvsControl_3dpi"]), n = Inf)$table

contrasts_smith <- makeContrasts(
  HPvsControl_1dpi = HP_1 - control_1,
  HPvsControl_3dpi = HP_3 - control_3,
  levels = design_smith
)
smith_1 <- topTags(glmLRT(fit_smith, contrast = contrasts_smith[, "HPvsControl_1dpi"]), n = Inf)$table
smith_3 <- topTags(glmLRT(fit_smith, contrast = contrasts_smith[, "HPvsControl_3dpi"]), n = Inf)$table

contrasts_morris <- makeContrasts(
  HPvsControl_2.2_2dpi   = HP_2.2_2   - control_0,
  HPvsControl_2.3.2.1_2dpi = HP_2.3.2.1_2 - control_0,
  levels = design_morris
)
morris_2.2_2     <- topTags(glmLRT(fit_morris, contrast = contrasts_morris[, "HPvsControl_2.2_2dpi"]),     n = Inf)$table
morris_2.3.2.1_2 <- topTags(glmLRT(fit_morris, contrast = contrasts_morris[, "HPvsControl_2.3.2.1_2dpi"]), n = Inf)$table
```

## 4.3 Summary of differentially expressed genes per study

```{r}
# Number of up/downregulated genes (FDR < 0.05) per study and timepoint
deg_summary <- function(study_name, dpi, df, fdr_col = "FDR", logfc_col = "logFC", threshold = 0.05) {
  data.frame(
    Study = study_name,
    DPI = dpi,
    Upregulated = sum(df[[fdr_col]] < threshold & df[[logfc_col]] > 0, na.rm = TRUE),
    Downregulated = sum(df[[fdr_col]] < threshold & df[[logfc_col]] < 0, na.rm = TRUE)
  )
}

deg_counts <- bind_rows(
  deg_summary("Campbell 2021", 1, campbell_1),
  deg_summary("Campbell 2021", 2, campbell_2),
  deg_summary("Campbell 2021", 3, campbell_3),
  deg_summary("Huang 2013",    1, huang_1),
  deg_summary("Huang 2013",    2, huang_2),
  deg_summary("Huang 2013",    3, huang_3),
  deg_summary("Smith 2015",    1, smith_1),
  deg_summary("Smith 2015",    3, smith_3),
  deg_summary("Morris 2023 (2.2.2.1)",  2, morris_2.2_2),
  deg_summary("Morris 2023 (2.3.2.1a)", 2, morris_2.3.2.1_2)
) %>%
  mutate(Total = Upregulated + Downregulated) %>%
  arrange(DPI, Study)

knitr::kable(deg_counts,
             caption = "Number of upregulated and downregulated genes (FDR < 0.05) per study and day post infection")

rm(deg_summary)
```

## 4.4 Venn diagrams

```{r}
# Helper: gene IDs passing a criterion
sig_genes <- function(df, col, threshold = 0.05, direction = NULL) {
  mask <- df[[col]] <= threshold
  if (!is.null(direction)) mask <- mask & (if (direction == "up") df$logFC > 0 else df$logFC < 0)
  rownames(df)[mask]
}

# P-value Venns
ggVennDiagram(list(
  `Morris 2.2 2dpi`   = sig_genes(morris_2.2_2,     "PValue"),
  `Morris 2.3.2.1 2dpi` = sig_genes(morris_2.3.2.1_2, "PValue"),
  `Campbell 2dpi`     = sig_genes(campbell_2,        "PValue"),
  `Huang 2dpi`        = sig_genes(huang_2,           "PValue")
)) + scale_fill_gradient(low = "white", high = "blue")

ggVennDiagram(list(
  `Campbell 1dpi` = sig_genes(campbell_1, "PValue"),
  `Huang 1dpi`    = sig_genes(huang_1,    "PValue"),
  `Smith 1dpi`    = sig_genes(smith_1,    "PValue")
)) + scale_fill_gradient(low = "white", high = "red")

ggVennDiagram(list(
  `Campbell 3dpi` = sig_genes(campbell_3, "PValue"),
  `Huang 3dpi`    = sig_genes(huang_3,    "PValue"),
  `Smith 3dpi`    = sig_genes(smith_3,    "PValue")
)) + scale_fill_gradient(low = "white", high = "purple")

# FDR Venns
ggVennDiagram(list(
  `Morris 2.2 2dpi`     = sig_genes(morris_2.2_2,     "FDR"),
  `Morris 2.3.2.1 2dpi` = sig_genes(morris_2.3.2.1_2, "FDR"),
  `Campbell 2dpi`       = sig_genes(campbell_2,        "FDR"),
  `Huang 2dpi`          = sig_genes(huang_2,           "FDR")
)) + scale_fill_gradient(low = "white", high = "blue")

ggVennDiagram(list(
  `Campbell 1dpi` = sig_genes(campbell_1, "FDR"),
  `Huang 1dpi`    = sig_genes(huang_1,    "FDR"),
  `Smith 1dpi`    = sig_genes(smith_1,    "FDR")
)) + scale_fill_gradient(low = "white", high = "red")

ggVennDiagram(list(
  `Campbell 3dpi` = sig_genes(campbell_3, "FDR"),
  `Huang 3dpi`    = sig_genes(huang_3,    "FDR"),
  `Smith 3dpi`    = sig_genes(smith_3,    "FDR")
)) + scale_fill_gradient(low = "white", high = "purple")

# FDR up-regulated Venns
ggVennDiagram(list(
  `Morris 2.2 2dpi`     = sig_genes(morris_2.2_2,     "FDR", direction = "up"),
  `Morris 2.3.2.1 2dpi` = sig_genes(morris_2.3.2.1_2, "FDR", direction = "up"),
  `Campbell 2dpi`       = sig_genes(campbell_2,        "FDR", direction = "up"),
  `Huang 2dpi`          = sig_genes(huang_2,           "FDR", direction = "up")
)) + scale_fill_gradient(low = "white", high = "blue")

ggVennDiagram(list(
  `Campbell 1dpi` = sig_genes(campbell_1, "FDR", direction = "up"),
  `Huang 1dpi`    = sig_genes(huang_1,    "FDR", direction = "up"),
  `Smith 1dpi`    = sig_genes(smith_1,    "FDR", direction = "up")
)) + scale_fill_gradient(low = "white", high = "red")

ggVennDiagram(list(
  `Campbell 3dpi` = sig_genes(campbell_3, "FDR", direction = "up"),
  `Huang 3dpi`    = sig_genes(huang_3,    "FDR", direction = "up"),
  `Smith 3dpi`    = sig_genes(smith_3,    "FDR", direction = "up")
)) + scale_fill_gradient(low = "white", high = "purple")

# FDR down-regulated Venns
ggVennDiagram(list(
  `Morris 2.2 2dpi`     = sig_genes(morris_2.2_2,     "FDR", direction = "down"),
  `Morris 2.3.2.1 2dpi` = sig_genes(morris_2.3.2.1_2, "FDR", direction = "down"),
  `Campbell 2dpi`       = sig_genes(campbell_2,        "FDR", direction = "down"),
  `Huang 2dpi`          = sig_genes(huang_2,           "FDR", direction = "down")
)) + scale_fill_gradient(low = "white", high = "blue")

ggVennDiagram(list(
  `Campbell 1dpi` = sig_genes(campbell_1, "FDR", direction = "down"),
  `Huang 1dpi`    = sig_genes(huang_1,    "FDR", direction = "down"),
  `Smith 1dpi`    = sig_genes(smith_1,    "FDR", direction = "down")
)) + scale_fill_gradient(low = "white", high = "red")

ggVennDiagram(list(
  `Campbell 3dpi` = sig_genes(campbell_3, "FDR", direction = "down"),
  `Huang 3dpi`    = sig_genes(huang_3,    "FDR", direction = "down"),
  `Smith 3dpi`    = sig_genes(smith_3,    "FDR", direction = "down")
)) + scale_fill_gradient(low = "white", high = "purple")

rm(sig_genes)
```

# 5. P-value combination and jackknife sensitivity analysis
## 5.1 P-value distributions under the null

```{r}
for (df_name in c("smith_1", "smith_3",
                  "campbell_1", "campbell_2", "campbell_3",
                  "morris_2.3.2.1_2", "morris_2.2_2",
                  "huang_1", "huang_2", "huang_3")) {
  hist(get(df_name)$PValue, 20, main = df_name, xlab = "P-value")
}
```

## 5.2 Meta-analysis function and p-value combination

```{r}
metaAnalysis <- function(study_list, study_names, weights) {
  common_genes <- Reduce(intersect, lapply(study_list, rownames))
  gene_name_map <- setNames(annotation_data$external_gene_name, annotation_data$ensembl_gene_id)

  out <- data.frame(gene = common_genes,
                    p_combined = NA_real_,
                    eff_direction = NA_character_,
                    same_direction = NA,
                    studies_included = NA_character_,
                    stringsAsFactors = FALSE)
  for (s in study_names) {
    out[[paste0("logFC_",  s)]] <- NA_real_
    out[[paste0("PValue_", s)]] <- NA_real_
    out[[paste0("FDR_",    s)]] <- NA_real_
  }

  for (i in seq_along(common_genes)) {
    gene <- common_genes[i]
    p_values <- fc_values <- numeric()
    studies  <- character()

    for (j in seq_along(study_list)) {
      df    <- study_list[[j]]
      study <- study_names[j]
      out[i, paste0("logFC_",  study)] <- df[gene, "logFC"]
      out[i, paste0("PValue_", study)] <- df[gene, "PValue"]
      out[i, paste0("FDR_",    study)] <- df[gene, "FDR"]
      p_values  <- c(p_values,  df[gene, "PValue"])
      fc_values <- c(fc_values, df[gene, "logFC"])
      studies   <- c(studies,   study)
    }

    wf <- wFisher(p = p_values, weight = weights,
                  is.onetail = FALSE, eff.sign = sign(fc_values))
    out$p_combined[i]     <- wf$p
    out$eff_direction[i]  <- wf$overall.eff.direction
    out$same_direction[i] <- (all(fc_values > 0) || all(fc_values < 0))
    out$studies_included[i] <- paste(studies, collapse = ", ")
  }

  out$gene_name    <- gene_name_map[out$gene]
  out$FDR_combined <- p.adjust(out$p_combined, method = "BH")
  col_order <- c("gene", "gene_name", "p_combined", "FDR_combined",
                 "same_direction", "studies_included",
                 setdiff(names(out), c("gene", "gene_name", "p_combined",
                                       "FDR_combined", "same_direction",
                                       "studies_included")))
  out[order(out$FDR_combined), col_order]
}

combined_1 <- metaAnalysis(
  study_list   = list(campbell_1, smith_1, huang_1),
  study_names  = c("campbell_1", "smith_1", "huang_1"),
  weights      = c(4, 3, 1)
)
combined_2 <- metaAnalysis(
  study_list   = list(morris_2.2_2, morris_2.3.2.1_2, campbell_2, huang_2),
  study_names  = c("morris_2.2_2", "morris_2.3.2.1_2", "campbell_2", "huang_2"),
  weights      = c(3, 3, 4, 1)
)
combined_3 <- metaAnalysis(
  study_list   = list(campbell_3, smith_3, huang_3),
  study_names  = c("campbell_3", "smith_3", "huang_3"),
  weights      = c(4, 3, 1)
)
print(head(combined_3))
```

## 5.3 Jackknife sampling

```{r}
jackknife_meta2 <- function(study_list, study_names, weights) {
  setNames(
    lapply(seq_along(study_list), function(i) {
      metaAnalysis(study_list[-i], study_names[-i], weights[-i])
    }),
    paste0("jackknife_excluding_", study_names)
  )
}

jackknife_results_1 <- jackknife_meta2(
  study_list  = list(campbell_1 = campbell_1, smith_1 = smith_1, huang_1 = huang_1),
  study_names = c("campbell_1", "smith_1", "huang_1"),
  weights     = c(4, 3, 1)
)

jackknife_results_2 <- jackknife_meta2(
  study_list  = list(morris_2.2_2 = morris_2.2_2, morris_2.3.2.1_2 = morris_2.3.2.1_2,
                     campbell_2 = campbell_2, huang_2 = huang_2),
  study_names = c("morris_2.2_2", "morris_2.3.2.1_2", "campbell_2", "huang_2"),
  weights     = c(3, 3, 4, 1)
)

jackknife_results_3 <- jackknife_meta2(
  study_list  = list(campbell_3 = campbell_3, smith_3 = smith_3, huang_3 = huang_3),
  study_names = c("campbell_3", "smith_3", "huang_3"),
  weights     = c(4, 3, 1)
)
```

## 5.4 Build jackknife flags

```{r}
# Helper to join jackknife results and compute flags
add_jackknife_flags_1_3 <- function(combined, jack_results, day,
                                    study_cols, day_suffix) {
  # Study-specific FDR column names in combined
  fdr_cols <- paste0("FDR_", study_cols, "_", day_suffix)

  combined %>%
    left_join(
      jack_results[[paste0("jackknife_excluding_campbell_", day)]] %>%
        dplyr::select(gene,
               p_combined_no_campbell   = p_combined,
               FDR_combined_no_campbell = FDR_combined,
               same_direction_no_campbell = same_direction,
               eff_direction_no_campbell  = eff_direction),
      by = "gene"
    ) %>%
    left_join(
      jack_results[[paste0("jackknife_excluding_smith_", day)]] %>%
        dplyr::select(gene,
               p_combined_no_smith   = p_combined,
               FDR_combined_no_smith = FDR_combined,
               same_direction_no_smith = same_direction,
               eff_direction_no_smith  = eff_direction),
      by = "gene"
    ) %>%
    left_join(
      jack_results[[paste0("jackknife_excluding_huang_", day)]] %>%
        dplyr::select(gene,
               p_combined_no_huang   = p_combined,
               FDR_combined_no_huang = FDR_combined,
               same_direction_no_huang = same_direction,
               eff_direction_no_huang  = eff_direction),
      by = "gene"
    ) %>%
    mutate(
      jackknife = as.integer(
        eff_direction_no_campbell == eff_direction_no_smith &
        eff_direction_no_campbell == eff_direction_no_huang &
        FDR_combined_no_campbell < 0.05 &
        FDR_combined_no_smith    < 0.05 &
        FDR_combined_no_huang    < 0.05
      ),
      combined = as.integer(FDR_combined < 0.05),
      campbell = as.integer(.data[[paste0("FDR_campbell_", day_suffix)]] < 0.05),
      smith    = as.integer(.data[[paste0("FDR_smith_",    day_suffix)]] < 0.05),
      huang    = as.integer(.data[[paste0("FDR_huang_",    day_suffix)]] < 0.05)
    )
}

combined_1_jackknife <- add_jackknife_flags_1_3(combined_1, jackknife_results_1,
                                                 day = 1, day_suffix = 1,
                                                 study_cols = c("campbell", "smith", "huang"))
combined_3_jackknife <- add_jackknife_flags_1_3(combined_3, jackknife_results_3,
                                                 day = 3, day_suffix = 3,
                                                 study_cols = c("campbell", "smith", "huang"))

# Day 2 has 4 studies (two Morris sub-strains)
combined_2_jackknife <- combined_2 %>%
  left_join(
    jackknife_results_2$jackknife_excluding_campbell_2 %>%
      dplyr::select(gene,
             p_combined_no_campbell   = p_combined,
             FDR_combined_no_campbell = FDR_combined,
             same_direction_no_campbell = same_direction,
             eff_direction_no_campbell  = eff_direction),
    by = "gene"
  ) %>%
  left_join(
    jackknife_results_2$jackknife_excluding_huang_2 %>%
      dplyr::select(gene,
             p_combined_no_huang   = p_combined,
             FDR_combined_no_huang = FDR_combined,
             same_direction_no_huang = same_direction,
             eff_direction_no_huang  = eff_direction),
    by = "gene"
  ) %>%
  left_join(
    jackknife_results_2$`jackknife_excluding_morris_2.3.2.1_2` %>%
      dplyr::select(gene,
             p_combined_no_morris_2_3_2_1   = p_combined,
             FDR_combined_no_morris_2_3_2_1 = FDR_combined,
             same_direction_no_morris_2_3_2_1 = same_direction,
             eff_direction_no_morris_2_3_2_1  = eff_direction),
    by = "gene"
  ) %>%
  left_join(
    jackknife_results_2$`jackknife_excluding_morris_2.2_2` %>%
      dplyr::select(gene,
             p_combined_no_morris_2_2   = p_combined,
             FDR_combined_no_morris_2_2 = FDR_combined,
             same_direction_no_morris_2_2 = same_direction,
             eff_direction_no_morris_2_2  = eff_direction),
    by = "gene"
  ) %>%
  mutate(
    jackknife = as.integer(
      eff_direction_no_campbell == eff_direction_no_huang &
      eff_direction_no_campbell == eff_direction_no_morris_2_3_2_1 &
      eff_direction_no_campbell == eff_direction_no_morris_2_2 &
      FDR_combined_no_campbell      < 0.05 &
      FDR_combined_no_huang         < 0.05 &
      FDR_combined_no_morris_2_3_2_1 < 0.05 &
      FDR_combined_no_morris_2_2    < 0.05
    ),
    combined       = as.integer(FDR_combined < 0.05),
    campbell       = as.integer(FDR_campbell_2     < 0.05),
    huang          = as.integer(FDR_huang_2         < 0.05),
    morris_2_3_2_1 = as.integer(`FDR_morris_2.3.2.1_2` < 0.05),
    morris_2_2     = as.integer(`FDR_morris_2.2_2`     < 0.05)
  )

rm(add_jackknife_flags_1_3)
```

## 5.5 UpSetR plots

```{r}
# Descriptive set labels used for all UpSet plots in this section and the
# next; original column names are restored at the end of section 5.6 so
# downstream filtering code (section 6+) keeps working
orig1 <- colnames(combined_1_jackknife)
orig2 <- colnames(combined_2_jackknife)
orig3 <- colnames(combined_3_jackknife)

rename_cols <- function(df, map) {
  hit <- names(df) %in% names(map)
  names(df)[hit] <- map[names(df)[hit]]
  df
}

study_labels <- c(
  jackknife = "Jackknife", combined = "Combined",
  smith = "Smith 2015", campbell = "Campbell 2021", huang = "Huang 2013",
  morris_2_2 = "Morris 2023 (2.2.2.1)", morris_2_3_2_1 = "Morris 2023 (2.3.2.1a)"
)

combined_1_jackknife <- rename_cols(combined_1_jackknife, study_labels)
combined_2_jackknife <- rename_cols(combined_2_jackknife, study_labels)
combined_3_jackknife <- rename_cols(combined_3_jackknife, study_labels)

upset(combined_1_jackknife, sets = c("Smith 2015", "Campbell 2021", "Huang 2013", "Combined"))
upset(combined_2_jackknife, sets = c("Huang 2013", "Campbell 2021", "Morris 2023 (2.2.2.1)", "Morris 2023 (2.3.2.1a)", "Combined"))
upset(combined_3_jackknife, sets = c("Smith 2015", "Campbell 2021", "Huang 2013", "Combined"))

upset(combined_1_jackknife, sets = c("Jackknife", "Smith 2015", "Campbell 2021", "Huang 2013"))
upset(combined_2_jackknife, sets = c("Jackknife", "Huang 2013", "Campbell 2021", "Morris 2023 (2.2.2.1)", "Morris 2023 (2.3.2.1a)"))
upset(combined_3_jackknife, sets = c("Jackknife", "Smith 2015", "Campbell 2021", "Huang 2013"))
```

## 5.6 UpSetR with combined flag (saved for publication)

```{r}
stopifnot(all(c("Jackknife","Smith 2015","Campbell 2021","Huang 2013","Combined") %in% names(combined_1_jackknife)))

# Print each plot into the knitted document, and additionally save a
# high-resolution PNG (300 dpi) and a vector PDF, suitable for journal
# submission
save_upset <- function(df, sets, basename, width_in = 10, height_in = 7) {
  print(upset(df, sets = sets))

  png(paste0(basename, ".png"), width = width_in, height = height_in, units = "in", res = 300)
  print(upset(df, sets = sets))
  dev.off()

  pdf(paste0(basename, ".pdf"), width = width_in, height = height_in)
  print(upset(df, sets = sets))
  dev.off()
}

save_upset(combined_1_jackknife,
           c("Jackknife", "Smith 2015", "Campbell 2021", "Huang 2013", "Combined"),
           "upset_combined_1")
save_upset(combined_2_jackknife,
           c("Jackknife", "Huang 2013", "Campbell 2021", "Morris 2023 (2.2.2.1)", "Morris 2023 (2.3.2.1a)", "Combined"),
           "upset_combined_2")
save_upset(combined_3_jackknife,
           c("Jackknife", "Smith 2015", "Campbell 2021", "Huang 2013", "Combined"),
           "upset_combined_3")

# Restore original column names so downstream code (sections 6+) keeps
# working with the original snake_case names
colnames(combined_1_jackknife) <- orig1
colnames(combined_2_jackknife) <- orig2
colnames(combined_3_jackknife) <- orig3
rm(orig1, orig2, orig3, rename_cols, study_labels, save_upset)
```

## 5.7 Jackknife-positive gene counts by direction

```{r}
# Number of jackknife-positive genes (FDR_combined < 0.05 and a consistent
# effect direction across every leave-one-study-out iteration), split by
# direction, per day
jackknife_direction_summary <- function(dpi, df) {
  data.frame(
    DPI = dpi,
    Upregulated   = sum(df$jackknife == 1 & df$eff_direction == "+", na.rm = TRUE),
    Downregulated = sum(df$jackknife == 1 & df$eff_direction == "-", na.rm = TRUE)
  )
}

jackknife_direction_counts <- bind_rows(
  jackknife_direction_summary(1, combined_1_jackknife),
  jackknife_direction_summary(2, combined_2_jackknife),
  jackknife_direction_summary(3, combined_3_jackknife)
) %>%
  mutate(Total = Upregulated + Downregulated)

knitr::kable(jackknife_direction_counts,
             caption = "Number of jackknife-positive genes (FDR < 0.05, consistent direction across leave-one-study-out iterations) by direction, per day post infection")

rm(jackknife_direction_summary)
```

# 6. Interactive datatables
## 6.0 Add gene descriptions to all combined tables

```{r}
# All datatables below (6.1-6.8) come from combined_X_jackknife. Those
# dataframes originally only carried "gene" (ensembl_gene_id) and
# "gene_name" (symbol) as identifier columns, so several of the raw tables
# effectively repeated the gene twice with no human-readable description.
# Add the annotation "description" once here, right after "gene", so every
# interactive table below shows it consistently as the second column.
add_description <- function(df) {
  df %>%
    left_join(annotation_data %>% dplyr::select(gene = ensembl_gene_id, description),
              by = "gene") %>%
    relocate(description, .after = gene)
}

combined_1_jackknife <- add_description(combined_1_jackknife)
combined_2_jackknife <- add_description(combined_2_jackknife)
combined_3_jackknife <- add_description(combined_3_jackknife)

rm(add_description)
```

## 6.1 Top 20 differentially expressed genes (jackknife) per day

```{r}
# Top 20 jackknife-supported DEGs per day, ranked by the combined FDR
top20_jackknife <- function(df) {
  df %>%
    filter(jackknife == 1) %>%
    arrange(FDR_combined) %>%
    slice_head(n = 20) %>%
    transmute(
      Gene = if_else(!is.na(gene_name) & gene_name != "", gene_name, gene),
      Description = description,
      FDR = signif(FDR_combined, 3),
      Direction = if_else(eff_direction == "+", "Up", "Down")
    )
}

datatable(top20_jackknife(combined_1_jackknife), rownames = FALSE,
          caption = "Top 20 jackknife-supported differentially expressed genes - 1 dpi")
datatable(top20_jackknife(combined_2_jackknife), rownames = FALSE,
          caption = "Top 20 jackknife-supported differentially expressed genes - 2 dpi")
datatable(top20_jackknife(combined_3_jackknife), rownames = FALSE,
          caption = "Top 20 jackknife-supported differentially expressed genes - 3 dpi")

rm(top20_jackknife)
```

## 6.2 Full tables

```{r, message=FALSE}
datatable(combined_1_jackknife)
datatable(combined_2_jackknife)
datatable(combined_3_jackknife)
```

## 6.3 Tables with genes found only by combined and not the different studies

```{r}
datatable(combined_1_jackknife %>% filter(combined == 1, smith != 1, huang != 1, campbell != 1))
datatable(combined_2_jackknife %>% filter(combined == 1, morris_2_2 != 1, morris_2_3_2_1 != 1, huang != 1, campbell != 1))
datatable(combined_3_jackknife %>% filter(combined == 1, smith != 1, huang != 1, campbell != 1))
```

## 6.4 Tables with genes found by combined

```{r}
datatable(combined_1_jackknife %>% filter(combined == 1))
datatable(combined_2_jackknife %>% filter(combined == 1))
datatable(combined_3_jackknife %>% filter(combined == 1))
```

## 6.5 Tables with genes found by all studies and combined

```{r}
datatable(combined_1_jackknife %>% filter(combined == 1, smith == 1, huang == 1, campbell == 1))
datatable(combined_2_jackknife %>% filter(combined == 1, morris_2_2 == 1, morris_2_3_2_1 == 1, huang == 1, campbell == 1))
datatable(combined_3_jackknife %>% filter(combined == 1, smith == 1, huang == 1, campbell == 1))
```

## 6.6 Tables with genes found by Jackknife and not the other studies

```{r}
datatable(combined_1_jackknife %>% filter(jackknife == 1, smith != 1, huang != 1, campbell != 1))
datatable(combined_2_jackknife %>% filter(jackknife == 1, morris_2_2 != 1, morris_2_3_2_1 != 1, huang != 1, campbell != 1))
datatable(combined_3_jackknife %>% filter(jackknife == 1, smith != 1, huang != 1, campbell != 1))
```

## 6.7 Tables with genes found by jackknife

```{r}
datatable(combined_1_jackknife %>% filter(jackknife == 1))
datatable(combined_2_jackknife %>% filter(jackknife == 1))
datatable(combined_3_jackknife %>% filter(jackknife == 1))
```

## 6.8 Tables with genes found by all studies and jackknife

```{r}
datatable(combined_1_jackknife %>% filter(jackknife == 1, smith == 1, huang == 1, campbell == 1))
datatable(combined_2_jackknife %>% filter(jackknife == 1, morris_2_2 == 1, morris_2_3_2_1 == 1, huang == 1, campbell == 1))
datatable(combined_3_jackknife %>% filter(jackknife == 1, smith == 1, huang == 1, campbell == 1))
```

## 6.9 Genes jackknife-positive at all three days (rankable by FDR)

```{r}
# Genes with jackknife == 1 at every timepoint. Each day's combined FDR and
# effect direction are kept as separate, clearly-named columns so the table
# can be re-ranked by clicking any of the three FDR_combined_dayN headers.
genes_all3_jackknife <- Reduce(intersect, list(
  combined_1_jackknife$gene[combined_1_jackknife$jackknife == 1],
  combined_2_jackknife$gene[combined_2_jackknife$jackknife == 1],
  combined_3_jackknife$gene[combined_3_jackknife$jackknife == 1]
))

build_day_summary <- function(day_num, df, genes) {
  df %>%
    filter(gene %in% genes) %>%
    dplyr::select(gene, gene_name, description, FDR_combined, eff_direction) %>%
    rename_with(~ paste0(., "_day", day_num), c(FDR_combined, eff_direction))
}

d1 <- build_day_summary(1, combined_1_jackknife, genes_all3_jackknife)
d2 <- build_day_summary(2, combined_2_jackknife, genes_all3_jackknife) %>%
  dplyr::select(-gene_name, -description)
d3 <- build_day_summary(3, combined_3_jackknife, genes_all3_jackknife) %>%
  dplyr::select(-gene_name, -description)

jackknife_all_days <- d1 %>%
  inner_join(d2, by = "gene") %>%
  inner_join(d3, by = "gene")

default_sort_col <- which(names(jackknife_all_days) == "FDR_combined_day1") - 1

datatable(jackknife_all_days, rownames = FALSE,
          caption = "Genes that are jackknife-positive at 1, 2, and 3 dpi - click any FDR_combined_dayN column header to re-rank",
          options = list(order = list(list(default_sort_col, "asc"))))

rm(d1, d2, d3, build_day_summary, genes_all3_jackknife, default_sort_col)
```

# 7. Log fold-change bar plots
## 7.1 Helper function and shared theme

```{r}
# Shared bar-plot helper to avoid repeating pivot+ggplot code
barplot_logfc <- function(df, gene_col, logfc_cols, title,
                          filter_expr = TRUE, n = 20) {
  df %>%
    filter(!is.na(gene_name) & gene_name != "", {{ filter_expr }}) %>%
    slice_head(n = n) %>%
    dplyr::select(all_of(c(gene_col, logfc_cols))) %>%
    pivot_longer(-all_of(gene_col), names_to = "source", values_to = "logFC") %>%
    ggplot(aes(x = .data[[gene_col]], y = logFC, fill = source)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title = title, x = "Gene", y = "logFC") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## 7.2 Top 20 combined genes

```{r}
barplot_logfc(combined_1_jackknife, "gene_name",
              c("logFC_campbell_1","logFC_smith_1","logFC_huang_1"),
              "Log Fold Changes for 1dpi",
              filter_expr = combined == 1)

barplot_logfc(combined_2_jackknife, "gene_name",
              c("logFC_morris_2.2_2","logFC_morris_2.3.2.1_2","logFC_campbell_2","logFC_huang_2"),
              "Log Fold Changes for 2dpi",
              filter_expr = combined == 1)

barplot_logfc(combined_3_jackknife, "gene_name",
              c("logFC_campbell_3","logFC_smith_3","logFC_huang_3"),
              "Log Fold Changes for 3dpi",
              filter_expr = combined == 1)
```

## 7.3 Top 20 common combined genes (in all studies)

```{r}
barplot_logfc(combined_1_jackknife, "gene_name",
              c("logFC_campbell_1","logFC_smith_1","logFC_huang_1"),
              "Log Fold Changes for 1dpi (all studies)",
              filter_expr = combined == 1 & campbell == 1 & smith == 1 & huang == 1)

barplot_logfc(combined_2_jackknife, "gene_name",
              c("logFC_morris_2.2_2","logFC_morris_2.3.2.1_2","logFC_campbell_2","logFC_huang_2"),
              "Log Fold Changes for 2dpi (all studies)",
              filter_expr = combined == 1 & campbell == 1 & morris_2_3_2_1 == 1 & morris_2_2 == 1 & huang == 1)

barplot_logfc(combined_3_jackknife, "gene_name",
              c("logFC_campbell_3","logFC_smith_3","logFC_huang_3"),
              "Log Fold Changes for 3dpi (all studies)",
              filter_expr = combined == 1 & campbell == 1 & smith == 1 & huang == 1)
```

## 7.4 Top 20 unique combined genes

```{r}
barplot_logfc(combined_1_jackknife, "gene_name",
              c("logFC_campbell_1","logFC_smith_1","logFC_huang_1"),
              "Log Fold Changes for 1dpi (unique to combined)",
              filter_expr = combined == 1 & campbell != 1 & smith != 1 & huang != 1)

barplot_logfc(combined_2_jackknife, "gene_name",
              c("logFC_morris_2.2_2","logFC_morris_2.3.2.1_2","logFC_campbell_2","logFC_huang_2"),
              "Log Fold Changes for 2dpi (unique to combined)",
              filter_expr = combined == 1 & campbell != 1 & morris_2_3_2_1 != 1 & morris_2_2 != 1 & huang != 1)

barplot_logfc(combined_3_jackknife, "gene_name",
              c("logFC_campbell_3","logFC_smith_3","logFC_huang_3"),
              "Log Fold Changes for 3dpi (unique to combined)",
              filter_expr = combined == 1 & campbell != 1 & smith != 1 & huang != 1)
```

## 7.5 Top 20 jackknife genes

```{r}
barplot_logfc(combined_1_jackknife, "gene_name",
              c("logFC_campbell_1","logFC_smith_1","logFC_huang_1"),
              "Log Fold Changes for 1dpi (jackknife)",
              filter_expr = jackknife == 1)

barplot_logfc(combined_2_jackknife, "gene_name",
              c("logFC_morris_2.2_2","logFC_morris_2.3.2.1_2","logFC_campbell_2","logFC_huang_2"),
              "Log Fold Changes for 2dpi (jackknife)",
              filter_expr = jackknife == 1)

barplot_logfc(combined_3_jackknife, "gene_name",
              c("logFC_campbell_3","logFC_smith_3","logFC_huang_3"),
              "Log Fold Changes for 3dpi (jackknife)",
              filter_expr = jackknife == 1)
```

## 7.6 Top 20 jackknife × common

```{r}
barplot_logfc(combined_1_jackknife, "gene_name",
              c("logFC_campbell_1","logFC_smith_1","logFC_huang_1"),
              "Log Fold Changes for 1dpi (jackknife, all studies)",
              filter_expr = jackknife == 1 & campbell == 1 & smith == 1 & huang == 1)

barplot_logfc(combined_2_jackknife, "gene_name",
              c("logFC_morris_2.2_2","logFC_morris_2.3.2.1_2","logFC_campbell_2","logFC_huang_2"),
              "Log Fold Changes for 2dpi (jackknife, all studies)",
              filter_expr = jackknife == 1 & campbell == 1 & morris_2_3_2_1 == 1 & morris_2_2 == 1 & huang == 1)

barplot_logfc(combined_3_jackknife, "gene_name",
              c("logFC_campbell_3","logFC_smith_3","logFC_huang_3"),
              "Log Fold Changes for 3dpi (jackknife, all studies)",
              filter_expr = jackknife == 1 & campbell == 1 & smith == 1 & huang == 1)
```

## 7.7 Top 20 jackknife unique genes

```{r}
barplot_logfc(combined_1_jackknife, "gene_name",
              c("logFC_campbell_1","logFC_smith_1","logFC_huang_1"),
              "Log Fold Changes for 1dpi (jackknife unique)",
              filter_expr = jackknife == 1 & campbell != 1 & smith != 1 & huang != 1)

barplot_logfc(combined_2_jackknife, "gene_name",
              c("logFC_morris_2.2_2","logFC_morris_2.3.2.1_2","logFC_campbell_2","logFC_huang_2"),
              "Log Fold Changes for 2dpi (jackknife unique)",
              filter_expr = jackknife == 1 & campbell != 1 & morris_2_3_2_1 != 1 & morris_2_2 != 1 & huang != 1)

barplot_logfc(combined_3_jackknife, "gene_name",
              c("logFC_campbell_3","logFC_smith_3","logFC_huang_3"),
              "Log Fold Changes for 3dpi (jackknife unique)",
              filter_expr = jackknife == 1 & campbell != 1 & smith != 1 & huang != 1)

rm(barplot_logfc)
```

## 7.8 Curated innate-immune gene panel

```{r}
# --- 7.8a Define the curated gene panel ---------------------------------
# `match_col` = "id"     -> match against ensembl_gene_id (used for genes with
#                            no reliable symbol in the duck annotation, so we
#                            match on the LOC identifier directly)
# `match_col` = "symbol" -> match against annotation_data$external_gene_name
# NOTE: TNFa's identifier is unconfirmed ("maybe LOC139998842" per the
# original request) - please verify this against the annotation before
# relying on that bar; a warning will also print at knit time if it doesn't
# resolve to anything.
gene_panel <- data.frame(
  label = c(
    "IL6", "IFNa2 (LOC101798847)", "IFNb (LOC101796052)", "IL8",
    "TNFa (LOC139998842)", "IL10", "IFNG", "RIGI", "MDA5 (IFIH1)",
    "LGP2 (DHX58)", "MAVS", "IRF7", "OAS (LOC101800895)", "TRIM25",
    "RNF125", "STING1", "TLR7", "TLR3", "TRIF (TICAM1)", "LSM14A",
    "RSAD2", "NLRR3 (LRRN3)", "PKR (EIF2AK2)", "CCL19",
    "MX (LOC101793492)", "IFITM3", "IFITM1"
  ),
  match_col = c(
    "symbol", "id", "id", "symbol",
    "id", "symbol", "symbol", "symbol", "symbol",
    "symbol", "symbol", "symbol", "id", "symbol",
    "symbol", "symbol", "symbol", "symbol", "symbol", "symbol",
    "symbol", "symbol", "symbol", "symbol",
    "id", "symbol", "symbol"
  ),
  match_value = c(
    "IL6", "LOC101798847", "LOC101796052", "IL8",
    "LOC139998842", "IL10", "IFNG", "RIGI", "IFIH1",
    "DHX58", "MAVS", "IRF7", "LOC101800895", "TRIM25",
    "RNF125", "STING1", "TLR7", "TLR3", "TICAM1", "LSM14A",
    "RSAD2", "LRRN3", "EIF2AK2", "CCL19",
    "LOC101793492", "IFITM3", "IFITM1"
  ),
  stringsAsFactors = FALSE
)
gene_panel$label <- factor(gene_panel$label, levels = gene_panel$label)

# Resolve each panel entry to its ensembl_gene_id in the annotation
resolve_panel_ids <- function(panel, annotation) {
  panel$ensembl_gene_id <- NA_character_
  for (i in seq_len(nrow(panel))) {
    hit <- if (panel$match_col[i] == "id") {
      annotation$ensembl_gene_id == panel$match_value[i]
    } else {
      !is.na(annotation$external_gene_name) & annotation$external_gene_name == panel$match_value[i]
    }
    ids <- unique(annotation$ensembl_gene_id[hit])
    if (length(ids) == 0) {
      warning("Gene panel: no annotation match for ", panel$label[i],
              " (", panel$match_col[i], " = ", panel$match_value[i], ")")
    } else {
      if (length(ids) > 1) {
        warning("Gene panel: multiple annotation matches for ", panel$label[i],
                "; using the first (", ids[1], ")")
      }
      panel$ensembl_gene_id[i] <- ids[1]
    }
  }
  panel
}

gene_panel <- resolve_panel_ids(gene_panel, annotation_data)
```

```{r}
# --- 7.8b Weighted "Combined" logFC + combined significance --------------
# metaAnalysis() (5.2) combines P-VALUES across studies via wFisher - it does
# not produce a combined effect size. logFC_combined here is a WEIGHTED MEAN
# of the per-study point estimates, using the exact same weights already
# chosen for wFisher (4/3/1 for Campbell/Smith/Huang at 1&3dpi; 3/3/4/1 for
# Morris 2.2/Morris 2.3.2.1/Campbell/Huang at 2dpi). This is a weighted
# average of point estimates, NOT a variance-weighted meta-analytic effect
# size (that would need a per-study SE, which isn't in the edgeR output).
#
# The Combined bar is flagged significant only when BOTH:
#   (a) FDR_combined < 0.05 (the existing `combined` flag), AND
#   (b) it is jackknife-positive, i.e. the `jackknife` flag already computed
#       in 5.4 (consistent direction and FDR_combined < 0.05 under every
#       leave-one-study-out re-analysis)
add_weighted_logfc <- function(df, study_cols, weights) {
  w <- weights / sum(weights)
  logfc_mat <- as.matrix(df[, study_cols])
  df$logFC_combined <- as.numeric(logfc_mat %*% w)
  df$combined_sig    <- as.integer(df$combined == 1 & df$jackknife == 1)
  df
}

combined_1_jackknife <- add_weighted_logfc(
  combined_1_jackknife,
  c("logFC_campbell_1", "logFC_smith_1", "logFC_huang_1"),
  c(4, 3, 1)
)
combined_2_jackknife <- add_weighted_logfc(
  combined_2_jackknife,
  c("logFC_morris_2.2_2", "logFC_morris_2.3.2.1_2", "logFC_campbell_2", "logFC_huang_2"),
  c(3, 3, 4, 1)
)
combined_3_jackknife <- add_weighted_logfc(
  combined_3_jackknife,
  c("logFC_campbell_3", "logFC_smith_3", "logFC_huang_3"),
  c(4, 3, 1)
)
```

```{r}
# --- 7.8c Plotting helpers -------------------------------------------------
# Colours: a fixed categorical palette (adjacent-pair colourblind-safe,
# validated for bar charts), one hue per study/estimate, so the same colour
# means the same study in every facet. "Combined" takes the next slot in the
# same sequence.
study_colors <- c(
  "Campbell 2021"          = "#2a78d6",  # slot 1 - blue
  "Huang 2013"              = "#eb6834",  # slot 2 - orange
  "Smith 2015"              = "#1baf7a",  # slot 3 - aqua
  "Morris 2023 (2.2.2.1)"   = "#eda100",  # slot 4 - yellow
  "Morris 2023 (2.3.2.1a)"  = "#e87ba4",  # slot 5 - magenta
  "Combined"                = "#008300"   # slot 6 - green
)

# Per-dpi study -> column map. Order here fixes both the legend order and
# the left-to-right bar order within each gene's group; "Combined" always
# comes last.
study_cols_all <- list(
  `1` = list(
    "Campbell 2021" = list(logfc = "logFC_campbell_1", sig = "campbell"),
    "Huang 2013"    = list(logfc = "logFC_huang_1",    sig = "huang"),
    "Smith 2015"    = list(logfc = "logFC_smith_1",    sig = "smith"),
    "Combined"      = list(logfc = "logFC_combined",   sig = "combined_sig")
  ),
  `2` = list(
    "Campbell 2021"          = list(logfc = "logFC_campbell_2",        sig = "campbell"),
    "Huang 2013"              = list(logfc = "logFC_huang_2",           sig = "huang"),
    "Morris 2023 (2.2.2.1)"   = list(logfc = "logFC_morris_2.2_2",      sig = "morris_2_2"),
    "Morris 2023 (2.3.2.1a)"  = list(logfc = "logFC_morris_2.3.2.1_2",  sig = "morris_2_3_2_1"),
    "Combined"                = list(logfc = "logFC_combined",          sig = "combined_sig")
  ),
  `3` = list(
    "Campbell 2021" = list(logfc = "logFC_campbell_3", sig = "campbell"),
    "Huang 2013"    = list(logfc = "logFC_huang_3",    sig = "huang"),
    "Smith 2015"    = list(logfc = "logFC_smith_3",    sig = "smith"),
    "Combined"      = list(logfc = "logFC_combined",   sig = "combined_sig")
  )
)

# Pivot the three per-dpi tables into one long data frame:
# gene x dpi x study, keeping only genes significant in at least one
# study/Combined at at least one dpi (same rule as before, applied once
# across the merged figure so all three facets share one gene ordering).
build_gene_panel_long <- function(df_list, panel, study_cols_all) {
  long <- purrr::imap_dfr(df_list, function(df, dpi_num) {
    cols <- study_cols_all[[dpi_num]]
    purrr::imap_dfr(cols, function(cc, study_label) {
      data.frame(
        ensembl_gene_id = df$gene,
        dpi_num         = dpi_num,
        dpi_label       = paste0(dpi_num, " dpi"),
        study           = study_label,
        logFC           = df[[cc$logfc]],
        significant     = df[[cc$sig]] == 1,
        stringsAsFactors = FALSE
      )
    })
  })

  long <- long %>%
    inner_join(panel %>% dplyr::select(ensembl_gene_id, label), by = "ensembl_gene_id") %>%
    filter(!is.na(logFC))

  keep_genes <- long %>%
    group_by(label) %>%
    summarise(any_sig = any(significant), .groups = "drop") %>%
    filter(any_sig) %>%
    pull(label)

  gene_levels_used <- levels(panel$label)[levels(panel$label) %in% as.character(keep_genes)]

  long %>%
    filter(label %in% keep_genes) %>%
    mutate(
      label     = factor(as.character(label), levels = gene_levels_used),
      dpi_label = factor(dpi_label, levels = c("1 dpi", "2 dpi", "3 dpi"))
    )
}

# One faceted bar chart (rows = dpi). Bars are manually positioned (rather
# than relying on ggplot's automatic dodge) so a non-significant bar can be
# marked with a precisely-placed "+" just beyond its tip, instead of
# being faded out - bars stay fully opaque and readable regardless of
# significance.
plot_gene_panel_faceted <- function(long, study_cols_all, study_colors,
                                    bar_width = 0.85, gap = 0.06,
                                    mark_offset = 0.35) {
  gene_levels <- levels(long$label)

  long <- long %>%
    mutate(
      gene_index = as.integer(label),
      study_n    = lengths(study_cols_all[dpi_num]),
      study_idx  = mapply(function(s, d) match(s, names(study_cols_all[[d]])),
                          as.character(study), dpi_num),
      slot_width = bar_width / study_n,
      x_center   = gene_index + (study_idx - (study_n + 1) / 2) * slot_width,
      xmin       = x_center - slot_width * (1 - gap) / 2,
      xmax       = x_center + slot_width * (1 - gap) / 2,
      # "+" sits just beyond the bar's tip - above for positive bars,
      # below for negative ones - rather than inside the bar
      mark_y     = ifelse(logFC >= 0, logFC + mark_offset, logFC - mark_offset),
      study      = factor(study, levels = names(study_colors))
    )

  mark_layer <- function(colour, size) {
    geom_text(data = function(d) dplyr::filter(d, !significant),
              aes(x = x_center, y = mark_y), label = "+",
              colour = colour, size = size, family = "sans", fontface = "bold")
  }

  ggplot(long) +
    geom_rect(aes(xmin = xmin, xmax = xmax, ymin = pmin(0, logFC), ymax = pmax(0, logFC),
                  fill = study),
              colour = "grey20", linewidth = 0.2) +
    mark_layer("grey15", 3.2) +
    geom_hline(yintercept = 0, colour = "#c3c2b7", linewidth = 0.4) +
    scale_fill_manual(values = study_colors, name = "Study / estimate", drop = FALSE) +
    scale_x_continuous(breaks = seq_along(gene_levels), labels = gene_levels,
                       expand = expansion(add = 0.6)) +
    # scales = "fixed" (the default) so all three dpi panels share one
    # y-axis range; axes = "all_y" forces every panel to draw its own
    # y-axis tick labels even though the scale is shared (ggplot2 >= 3.5.0
    # - upgrade ggplot2 if this argument errors on an older install)
    facet_wrap(~ dpi_label, ncol = 1, scales = "fixed", axes = "all_y") +
    labs(x = "Gene", y = expression(log[2]*"FC"),
         caption = "+ = not significant (per-study: FDR ≥ 0.05; Combined: FDR ≥ 0.05 and/or not jackknife-positive)") +
    theme_minimal(base_size = 12) +
    theme(
      axis.text.x        = element_text(angle = 45, hjust = 1),
      axis.text.y        = element_text(size = 9),
      strip.text         = element_text(face = "bold", size = 12, hjust = 0),
      strip.background   = element_blank(),
      panel.grid.minor   = element_blank(),
      panel.grid.major.x = element_blank(),
      plot.caption        = element_text(hjust = 0, size = 9, colour = "#52514e"),
      legend.position     = "bottom"
    )
}
```

```{r, fig.width=13, fig.height=15}
# --- 7.8d Build and render the combined figure -----------------------
# Replaces the previous three separate 1/2/3 dpi panels with one figure,
# faceted by dpi (clearly labelled per-facet - no overall title needed),
# including a weighted "Combined" bar per gene. Printed into the knitted
# HTML and saved as high-resolution PNG + vector PDF, same convention as
# the rest of section 7.
gene_panel_long <- build_gene_panel_long(
  list(`1` = combined_1_jackknife, `2` = combined_2_jackknife, `3` = combined_3_jackknife),
  gene_panel, study_cols_all
)

p_gene_panel_all <- plot_gene_panel_faceted(gene_panel_long, study_cols_all, study_colors)
print(p_gene_panel_all)

ggsave("gene_panel_all_dpi.png", p_gene_panel_all, width = 13, height = 15, dpi = 300)
ggsave("gene_panel_all_dpi.pdf", p_gene_panel_all, width = 13, height = 15)

rm(resolve_panel_ids, add_weighted_logfc, study_colors, study_cols_all,
   build_gene_panel_long, plot_gene_panel_faceted, gene_panel_long,
   gene_panel, p_gene_panel_all)
```

# 8. Gene Ontology (overrepresentation analysis)
## 8.1 Annotation and helper

```{r}
# GO annotation from NCBI gene2go (replaces BioMart)
g2g <- read.delim("duck_gene2go.tsv", quote = "") %>%
  filter(!grepl("NOT", Qualifier))

# Map GeneID -> symbol, so keys match rownames(counts) / df$gene
entrez2sym <- setNames(annotation_data$ensembl_gene_id,
                       as.character(annotation_data$entrezgene_id))

go_tbl <- g2g %>%
  mutate(gene = entrez2sym[as.character(GeneID)]) %>%
  filter(!is.na(gene)) %>%
  mutate(namespace_1003 = recode(Category,
                                 "Process"   = "biological_process",
                                 "Function"  = "molecular_function",
                                 "Component" = "cellular_component")) %>%
  distinct()

TERM2GENE <- go_tbl %>% dplyr::select(go_id = GO_ID, gene) %>% distinct()
TERM2NAME <- go_tbl %>% dplyr::select(go_id = GO_ID, name = GO_term) %>% distinct()

# Background: genes actually tested in the meta-analysis
plotgo <- function(target_genes, title = "GO Enrichment", universe = NULL) {
  go_enrich <- enricher(gene = target_genes, TERM2GENE = TERM2GENE,
                        TERM2NAME = TERM2NAME, universe = universe,
                        pvalueCutoff = 0.05)
  if (!is.null(go_enrich) && nrow(go_enrich@result) > 0)
    print(barplot(go_enrich, showCategory = 20, title = title))
  else
    message("No enrichment found.")
}
```

## 8.2 P-value combination genes

```{r, fig.height=15, fig.width=15}
for (day in 1:3) {
  df <- get(paste0("combined_", day, "_jackknife"))
  uni <- intersect(TERM2GENE$gene, df$gene)
  goi <- df %>% filter(combined == 1) %>% pull(gene) %>% unique() %>% na.omit()
  plotgo(goi, title = sprintf("DEGs after p-value combination, %d dpi", day),universe = uni)
  goi_up   <- df %>% filter(combined == 1, eff_direction == '+') %>% pull(gene) %>% unique() %>% na.omit()
  plotgo(goi_up, title = sprintf("Upregulated after p-value combination, %d dpi", day),universe = uni)
  goi_down <- df %>% filter(combined == 1, eff_direction == '-') %>% pull(gene) %>% unique() %>% na.omit()
  plotgo(goi_down, title = sprintf("Downregulated after p-value combination, %d dpi", day),universe = uni)
}
```

## 8.3 Jackknife genes

```{r, fig.height=15, fig.width=15}
for (day in 1:3) {
  df <- get(paste0("combined_", day, "_jackknife"))
  uni <- intersect(TERM2GENE$gene, df$gene)
  goi_up   <- df %>% filter(jackknife == 1, eff_direction == '+') %>% pull(gene) %>% unique() %>% na.omit()
  plotgo(goi_up,   title = sprintf("Upregulated after jackknife, %d dpi", day),universe = uni)
  goi_down <- df %>% filter(jackknife == 1, eff_direction == '-') %>% pull(gene) %>% unique() %>% na.omit()
  plotgo(goi_down, title = sprintf("Downregulated after jackknife, %d dpi", day),universe = uni)
}
```

# 9. KEGG enrichment
## 9.1 apla

```{r}
# Duck-native KEGG: NCBI GeneIDs from the T2T RefSeq annotation map directly
# onto KEGG organism 'apla' — no ortholog projection needed.
apla_ids <- sub("^ncbi-geneid:", "", KEGGREST::keggConv("ncbi-geneid", "apla"))

add_entrez <- function(df) {
  df %>%
    left_join(annotation_data %>% dplyr::select(gene = ensembl_gene_id, entrezgene_id),
              by = "gene") %>%
    filter(!is.na(entrezgene_id)) %>%
    mutate(entrezgene_id = as.character(entrezgene_id))
}

combined_1_jackknife_apla <- add_entrez(combined_1_jackknife)
combined_2_jackknife_apla <- add_entrez(combined_2_jackknife)
combined_3_jackknife_apla <- add_entrez(combined_3_jackknife)

plotkegg <- function(target_genes, title, organism, universe = NULL) {
  kegg_enrich <- enrichKEGG(gene = target_genes, organism = organism,
                            universe = universe, pvalueCutoff = 0.05)
  if (!is.null(kegg_enrich) && nrow(kegg_enrich@result) > 0)
    print(barplot(kegg_enrich, showCategory = 20, title = title))
  else
    message("No KEGG enrichment found.")
}
```

## 9.2 KEGG barplots

```{r, fig.width=15, fig.height=15}
kegg_subsets <- list(
  list(filter_expr = quote(combined == 1 & eff_direction == '+'), label = "upregulated (combined)"),
  list(filter_expr = quote(combined == 1 & eff_direction == '-'), label = "downregulated (combined)"),
  list(filter_expr = quote(jackknife == 1 & eff_direction == '+'), label = "upregulated (jackknife)"),
  list(filter_expr = quote(jackknife == 1 & eff_direction == '-'), label = "downregulated (jackknife)")
)

for (day in 1:3) {
  df  <- get(paste0("combined_", day, "_jackknife_apla"))
  uni <- intersect(apla_ids, unique(df$entrezgene_id))
  for (ss in kegg_subsets) {
    goi <- df %>% filter(!!ss$filter_expr) %>% pull(entrezgene_id) %>% unique() %>% na.omit()
    goi <- intersect(goi, uni)
    plotkegg(goi, sprintf("%s, %d dpi", ss$label, day), "apla", universe = uni)
  }
}
rm(kegg_subsets)
```

### 9.2.1 Common genes (jackknife, all studies)

```{r, fig.width=15, fig.height=15}
for (day in 1:3) {
  df  <- get(paste0("combined_", day, "_jackknife_apla"))
  uni <- intersect(apla_ids, unique(df$entrezgene_id))
  filter_expr <- if (day == 2)
    quote(jackknife == 1 & huang == 1 & campbell == 1 & morris_2_3_2_1 == 1 & morris_2_2 == 1)
  else
    quote(jackknife == 1 & huang == 1 & campbell == 1 & smith == 1)
  goi <- df %>% filter(!!filter_expr) %>% pull(entrezgene_id) %>% unique() %>% na.omit()
  goi <- intersect(goi, uni)
  plotkegg(goi, sprintf("common genes across all studies (jackknife), %d dpi", day), "apla",
           universe = uni)
}
```

# 10. Pathway visualisation (pathview)

```{r, message=FALSE, warning=FALSE}
for (day in 1:3) {
  obj_name <- paste0("combined_", day, "_jackknife_apla")
  df <- get(obj_name)
  assign(obj_name, df %>% mutate(direction_value = ifelse(eff_direction == "+", 1, -1)))
}

pathway_ids <- c(
  'apla05164', # Influenza A
  'apla04620', # Toll-like receptor
  'apla04621', # NOD-like receptor
  'apla04622', # RIG-I-like receptor
  'apla04210', # Apoptosis
  'apla04217'  # Necroptosis
)

# Confirm these maps exist for apla before looping
apla_paths <- names(KEGGREST::keggList("pathway", "apla"))
setdiff(pathway_ids, sub("^path:", "", apla_paths))   # expect character(0)

df_list <- list(
  day1 = combined_1_jackknife_apla,
  day2 = combined_2_jackknife_apla,
  day3 = combined_3_jackknife_apla
)

for (day_label in names(df_list)) {
  df <- df_list[[day_label]]
  for (subset_type in c("jackknife", "combined")) {
    if (!subset_type %in% colnames(df)) next
    subset_df <- df[df[[subset_type]] == 1, c("entrezgene_id", "direction_value")]
    if (nrow(subset_df) == 0) next
    gene_data <- setNames(subset_df$direction_value, as.character(subset_df$entrezgene_id))
    for (pathway in pathway_ids) {
      tryCatch(
        pathview(gene.data = gene_data, pathway.id = pathway, species = "apla",
                 gene.idtype = "KEGG",
                 verbose = TRUE, kegg.native = TRUE,
                 out.suffix = paste0(day_label, "_", subset_type),
                 limit = list(gene = 2, cpd = 1),
                 low = list(gene = "red", cpd = "blue"),
                 mid = list(gene = "gray", cpd = "gray"),
                 high = list(gene = "green", cpd = "yellow")),
        error = function(e) {
          message("pathview failed for ", pathway, " (", day_label, "/", subset_type, "): ", e$message)
          pathview(gene.data = gene_data, pathway.id = pathway, species = "apla",
                   gene.idtype = "KEGG",
                   verbose = TRUE, kegg.native = FALSE,
                   out.suffix = paste0(day_label, "_", subset_type, "_F"),
                   limit = list(gene = 2, cpd = 1),
                   low = list(gene = "red", cpd = "blue"),
                   mid = list(gene = "gray", cpd = "gray"),
                   high = list(gene = "green", cpd = "yellow"))
        }
      )
    }
  }
}
```

# 11. Comparison of KEGG pathways across studies (compareCluster)

```{r}
cluster_cols <- list(
  day1 = c("jackknife", "huang", "campbell", "smith"),
  day2 = c("jackknife", "huang", "campbell", "morris_2_3_2_1", "morris_2_2"),
  day3 = c("jackknife", "huang", "campbell", "smith")
)

long_title_base <- str_wrap(
  "Overrepresented KEGG Pathways In Differentially Expressed Genes Across Days Post Infection",
  width = 60
)

green_scale <- scale_fill_gradient(
  low  = rgb(188, 207, 0, maxColorValue = 255),
  high = rgb(58, 170, 53, maxColorValue = 255)
)

title_theme <- theme(plot.title = element_text(hjust = 0.5, size = 14, margin = margin(b = 10)))

run_kegg_compare <- function(df_list, cluster_cols, direction_filter = NULL, title_suffix = "") {
  for (day in names(df_list)) {
    df <- df_list[[day]]
    uni <- intersect(apla_ids, unique(as.character(df$entrezgene_id)))
    if (!is.null(direction_filter)) df <- df %>% filter(eff_direction == direction_filter)
    cols <- cluster_cols[[day]]
    geneclusters <- Filter(function(x) length(x) > 0, setNames(
      lapply(cols, function(col) {
        if (col %in% colnames(df))
          intersect(as.character(df$entrezgene_id[df[[col]] == 1]), uni) else NULL
      }), cols))
    if (length(geneclusters) == 0) { message(sprintf("No clusters for %s; skipping.", day)); next }
    cc <- compareCluster(geneclusters, fun = "enrichKEGG", organism = "apla",
                         universe = uni, pvalueCutoff = 0.05)
    long_title <- str_wrap(paste0(long_title_base, title_suffix), width = 60)
    print(dotplot(cc) + ggtitle(long_title) + green_scale + title_theme)
  }
}

run_kegg_compare(df_list, cluster_cols)
run_kegg_compare(df_list, cluster_cols, direction_filter = "+", title_suffix = " (Upregulated)")
run_kegg_compare(df_list, cluster_cols, direction_filter = "-", title_suffix = " (Downregulated)")
```

## 11.4 Jackknife genes across days (single compareCluster)

```{r}
uni_kegg_all <- Reduce(intersect, lapply(df_list, function(df)
  intersect(apla_ids, unique(as.character(df$entrezgene_id)))))

geneclusters_days <- setNames(
  lapply(names(df_list), function(day) {
    df <- df_list[[day]]
    intersect(unique(as.character(df$entrezgene_id[df$jackknife == 1 & !is.na(df$entrezgene_id)])),
              uni_kegg_all)
  }),
  c("Day 1", "Day 2", "Day 3")
)
geneclusters_days <- Filter(function(x) length(x) > 0, geneclusters_days)

cc_days <- compareCluster(geneclusters_days, fun = "enrichKEGG",
                          organism = "apla", universe = uni_kegg_all, pvalueCutoff = 0.05)
cc_days@compareClusterResult$Cluster <- factor(cc_days@compareClusterResult$Cluster,
                                               levels = names(geneclusters_days))
print(dotplot(cc_days) +
  ggtitle(str_wrap("KEGG Pathways in Jackknife-Positive Genes Across Days Post Infection", 60)) +
  green_scale + title_theme)

geneclusters6 <- unlist(lapply(seq_along(df_list), function(i) {
  df <- df_list[[i]]
  list(
    up   = intersect(unique(as.character(df$entrezgene_id[df$jackknife == 1 & df$eff_direction == '+'])), uni_kegg_all),
    down = intersect(unique(as.character(df$entrezgene_id[df$jackknife == 1 & df$eff_direction == '-'])), uni_kegg_all)
  ) |> setNames(paste0("Day ", i, c(" (Up)", " (Down)")))
}), recursive = FALSE)
geneclusters6 <- Filter(function(x) length(x) > 0, geneclusters6)

cc6 <- compareCluster(geneclusters6, fun = "enrichKEGG", organism = "apla",
                      universe = uni_kegg_all, pvalueCutoff = 0.05)
cc6@compareClusterResult$Cluster <- factor(cc6@compareClusterResult$Cluster, levels = names(geneclusters6))
print(dotplot(cc6) +
  ggtitle(str_wrap("KEGG Pathways in Jackknife-Positive Genes (Up/Down) Across Days", 60)) +
  green_scale + title_theme)
```

## 11.5 GO terms across days (custom GMT, enricher)

```{r}
# Adapter: gene2go table in the column names run_cc_namespace expects
go_ann <- go_tbl %>%
  dplyr::select(go_id = GO_ID, ensembl_gene_id = gene,
                name_1006 = GO_term, namespace_1003)

df_list_go <- list(
  day1 = combined_1_jackknife,
  day2 = combined_2_jackknife,
  day3 = combined_3_jackknife
)

uni_go_all <- Reduce(intersect, lapply(df_list_go, function(df) df$gene))

geneclusters_go <- setNames(
  lapply(df_list_go, function(df) {
    intersect(unique(as.character(na.omit(df$gene[df$jackknife == 1]))), uni_go_all)
  }),
  c("Day 1", "Day 2", "Day 3")
)
geneclusters_go <- Filter(function(x) length(x) > 0, geneclusters_go)

run_cc_namespace <- function(annotation_data, geneclusters, namespace_value,
                             pretty_ns = NULL, universe = NULL, pvalueCutoff = 0.05,
                             qvalueCutoff = 0.20, minGSSize = 5) {
  stopifnot("namespace_1003" %in% colnames(annotation_data))
  if (is.null(pretty_ns)) pretty_ns <- tools::toTitleCase(gsub("_", " ", namespace_value))

  ann_ns <- annotation_data %>%
    dplyr::filter(!is.na(go_id), !is.na(ensembl_gene_id), namespace_1003 == namespace_value)
  if (nrow(ann_ns) == 0) { message("No rows for namespace '", namespace_value, "'"); return(invisible(NULL)) }

  TERM2GENE_ns <- ann_ns %>% dplyr::select(go_id, ensembl_gene_id) %>% distinct()
  TERM2NAME_ns <- ann_ns %>% dplyr::select(go_id, name_1006) %>% distinct()

  bg <- if (is.null(universe)) unique(TERM2GENE_ns$ensembl_gene_id)
        else intersect(unique(TERM2GENE_ns$ensembl_gene_id), universe)

  cc_obj <- tryCatch(
    compareCluster(geneClusters = geneclusters, fun = "enricher",
                   TERM2GENE = TERM2GENE_ns, TERM2NAME = TERM2NAME_ns,
                   universe = bg, pAdjustMethod = "BH",
                   pvalueCutoff = pvalueCutoff, qvalueCutoff = qvalueCutoff,
                   minGSSize = minGSSize),
    error = function(e) { message("compareCluster error: ", e$message); NULL }
  )

  cc_df <- tryCatch(as.data.frame(cc_obj), error = function(e) NULL)
  if (is.null(cc_df) || nrow(cc_df) == 0) {
    message("No enrichment found for namespace '", namespace_value, "'")
    return(invisible(NULL))
  }

  cc_obj@compareClusterResult$Cluster <- factor(cc_obj@compareClusterResult$Cluster,
                                                 levels = names(geneclusters))
  p <- dotplot(cc_obj) +
    ggtitle(str_wrap(sprintf("GO %s — Jackknife Genes Across Days", pretty_ns), 60)) +
    green_scale + title_theme
  print(p)
  invisible(list(cc_object = cc_obj, cc_df = cc_df, plot = p))
}

ns_map <- c("biological_process" = "Biological Process",
            "molecular_function" = "Molecular Function",
            "cellular_component" = "Cellular Component")

res_bp    <- run_cc_namespace(go_ann, geneclusters_go, "biological_process", ns_map[1], universe = uni_go_all)
res_mf    <- run_cc_namespace(go_ann, geneclusters_go, "molecular_function", ns_map[2], universe = uni_go_all)
res_cc_go <- run_cc_namespace(go_ann, geneclusters_go, "cellular_component", ns_map[3], universe = uni_go_all)
```

## 11.6 Common genes across all days

```{r}
genes_all_jackknife <- Reduce(intersect, lapply(1:3, function(d)
  get(paste0("combined_", d, "_jackknife_apla")) %>%
    filter(jackknife == 1) %>% pull(gene)
))

build_day_df <- function(day, genes) {
  get(paste0("combined_", day, "_jackknife_apla")) %>%
    filter(gene %in% genes) %>%
    dplyr::select(gene, gene_name, p_combined, FDR_combined, eff_direction) %>%
    rename_with(~ paste0(., "_day", day), c(p_combined, FDR_combined, eff_direction))
}

d1 <- build_day_df(1, genes_all_jackknife)
d2 <- build_day_df(2, genes_all_jackknife) %>% dplyr::select(-gene_name)
d3 <- build_day_df(3, genes_all_jackknife) %>% dplyr::select(-gene_name)

merged_jackknife <- d1 %>%
  inner_join(d2, by = "gene") %>%
  inner_join(d3, by = "gene") %>%
  relocate(gene_name, .after = gene) %>%
  filter(!is.na(gene_name) & gene_name != "")

datatable(merged_jackknife, rownames = FALSE)
rm(d1, d2, d3, build_day_df, genes_all_jackknife)
```

# 12. Publication figure — KEGG + GO panel

```{r final_enrichment_figure, warning=FALSE, message=FALSE, fig.width=15, fig.height=30}
suppressPackageStartupMessages({
  library(enrichplot)
  library(patchwork)
})

geneclusters_kegg <- setNames(
  lapply(1:3, function(d) {
    df <- get(paste0("combined_", d, "_jackknife_apla"))
    intersect(unique(na.omit(as.character(df$entrezgene_id[df$jackknife == 1]))), uni_kegg_all)
  }),
  c("Day_1", "Day_2", "Day_3")
)
geneclusters_kegg <- Filter(function(x) length(x) > 0, geneclusters_kegg)

cc_kegg <- compareCluster(geneclusters_kegg, fun = "enrichKEGG",
                          organism = "apla", universe = uni_kegg_all, pvalueCutoff = 0.05)
cc_kegg@compareClusterResult$Cluster <- factor(cc_kegg@compareClusterResult$Cluster,
                                               levels = names(geneclusters_kegg))

ann_bp <- go_ann %>% filter(namespace_1003 == "biological_process")
TERM2GENE_BP <- ann_bp %>% dplyr::select(go_id, ensembl_gene_id) %>% distinct()
TERM2NAME_BP <- ann_bp %>% dplyr::select(go_id, name_1006) %>% distinct()

geneclusters_gobp <- setNames(
  lapply(1:3, function(d) {
    df <- get(paste0("combined_", d, "_jackknife"))
    intersect(unique(na.omit(as.character(df$gene[df$jackknife == 1]))), uni_go_all)
  }),
  c("Day_1", "Day_2", "Day_3")
)
geneclusters_gobp <- Filter(function(x) length(x) > 0, geneclusters_gobp)

cc_go_bp <- compareCluster(
  geneclusters_gobp, fun = "enricher",
  TERM2GENE = TERM2GENE_BP, TERM2NAME = TERM2NAME_BP,
  universe = intersect(unique(TERM2GENE_BP$ensembl_gene_id), uni_go_all),
  pAdjustMethod = "BH", pvalueCutoff = 0.05,
  qvalueCutoff = 0.20, minGSSize = 5
)
cc_go_bp@compareClusterResult$Cluster <- factor(cc_go_bp@compareClusterResult$Cluster,
                                                levels = names(geneclusters_gobp))
# Shared scale helpers
extract_minimal <- function(cc_obj) {
  df <- as.data.frame(cc_obj)
  df$GeneRatio_num <- sapply(df$GeneRatio, function(x) {
    parts <- strsplit(as.character(x), "/")[[1]]
    if (length(parts) != 2) return(NA_real_)
    as.numeric(parts[1]) / as.numeric(parts[2])
  })
  dplyr::select(df, p.adjust, GeneRatio = GeneRatio_num)
}

safe_range <- function(x, eps = 1e-6) {
  x <- x[is.finite(x)]
  if (length(x) == 0) return(c(0, 1))
  mn <- min(x); mx <- max(x)
  if (isTRUE(all.equal(mn, mx))) return(c(mn * (1 - eps), mx * (1 + eps)))
  c(mn, mx)
}

all_min   <- bind_rows(extract_minimal(cc_kegg), extract_minimal(cc_go_bp)) %>%
             filter(!is.na(p.adjust), !is.na(GeneRatio))
pval_range <- safe_range(all_min$p.adjust)
gr_range   <- safe_range(all_min$GeneRatio)

pub_fill <- scale_fill_gradient(
  low = rgb(188, 207, 0, maxColorValue = 255),
  high = rgb(58, 170, 53, maxColorValue = 255),
  limits = pval_range, name = "Adjusted p-value", na.value = "grey50"
)
pub_size <- scale_size(limits = gr_range, range = c(2, 8), name = "Gene Ratio")

p_kegg_pub <- dotplot(cc_kegg) +
  ggtitle("KEGG Pathways") + pub_fill + pub_size +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
        axis.text  = element_text(size = 9), legend.position = "right")

p_go_bp_pub <- dotplot(cc_go_bp) +
  ggtitle("GO Biological Processes") + pub_fill + pub_size +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
        axis.text  = element_text(size = 9), legend.position = "none")

final_pub_plot <- (p_kegg_pub | p_go_bp_pub) +
  plot_layout(guides = "collect", widths = c(1, 1)) +
  plot_annotation(
    title = str_wrap("Functional Enrichment of Jackknife-Positive Genes Across Days Post Infection", 80),
    tag_levels = "A",
    theme = theme(
      plot.title = element_text(hjust = 0.5, size = 16, face = "bold", margin = margin(b = 12)),
      plot.tag   = element_text(size = 14, face = "bold")
    )
  )

print(final_pub_plot)
ggsave("Figure_Functional_Enrichment_Jackknife.pdf", final_pub_plot,
       width = 12, height = 6, device = cairo_pdf)
ggsave("Figure_Functional_Enrichment_Jackknife.png", final_pub_plot,
       width = 12, height = 6, dpi = 600)

message("Figure exported: Figure_Functional_Enrichment_Jackknife.{pdf,png}")
```
