---
title: "meta_analysis"
author: "Killian"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Preparation
## 1.1 Setting seed for reproducibility

```{r}
set.seed(42)
```

## 1.2 Loading packages

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(ggplot2)
library(edgeR)
library(biomaRt)
library(metapro)
library(ggVennDiagram)
library(UpSetR)
library(DT)
library(clusterProfiler)
library(pathview)
library(gage)
library(gageData)
library(grid)
library(GO.db)
library(patchwork)
library(scales)
```

# 2. Preparation of the data
## 2.1 Loading count data and metadata

```{r}
# Read pre-processed counts and metadata
counts   <- read.csv("counts.csv",   row.names = 1)
metadata <- read.csv("metadata.csv", row.names = 1)

# Order counts to match metadata
counts <- counts[, metadata$Run]

# Split counts by study
campbell_counts <- counts[, metadata$Run[metadata$study == "campbell"], drop = FALSE]
huang_counts    <- counts[, metadata$Run[metadata$study == "huang"],    drop = FALSE]
smith_counts    <- counts[, metadata$Run[metadata$study == "smith"],    drop = FALSE]
morris_counts   <- counts[, metadata$Run[metadata$study == "morris"],   drop = FALSE]
```

## 2.2 Loading annotation

```{r}
# Load BioMart annotation for Anas platyrhynchos (downloaded from Ensembl)
annotation_data <- read.csv("Anas_platyrhynchos2_annotation.csv")

# Keep only relevant columns and unique rows
annotation_data <- annotation_data[-c(2, 3, 5, 6, 7)]
annotation_data <- distinct(annotation_data)

# Restrict to genes present in the count matrix and align row order
annotation_data <- annotation_data[annotation_data$ensembl_gene_id %in% rownames(counts), ]
annotation_data <- annotation_data[match(rownames(counts), annotation_data$ensembl_gene_id), ]

stopifnot(all(rownames(counts) == annotation_data$gene_id))
```

## 2.3 Creating DGEList objects

```{r}
# Ensure metadata matches counts columns exactly
metadata <- metadata %>%
  filter(Run %in% colnames(counts)) %>%
  .[match(colnames(counts), .$Run), ] %>%
  mutate(sample_id = paste(dpi, treatment, biological_replicate, sep = "_"))

# Full DGEList for initial QC
y <- DGEList(counts = counts, samples = metadata)

# Per-study DGELists
campbell <- DGEList(counts = campbell_counts, samples = metadata %>% filter(study == "campbell"), genes = annotation_data)
huang    <- DGEList(counts = huang_counts,    samples = metadata %>% filter(study == "huang"),    genes = annotation_data)
morris   <- DGEList(counts = morris_counts,   samples = metadata %>% filter(study == "morris"),   genes = annotation_data)
smith    <- DGEList(counts = smith_counts,    samples = metadata %>% filter(study == "smith"),    genes = annotation_data)

rm(campbell_counts, huang_counts, morris_counts, smith_counts)
```

# 3. Visualisation, filtering, normalisation
## 3.1 Raw data PCA

```{r, message=FALSE, warning=FALSE}
# Log-CPM PCA on all samples
pca_res <- prcomp(t(cpm(y, log = TRUE)), scale. = TRUE)
pca <- merge(
  transform(as.data.frame(pca_res$x), Run = rownames(pca_res$x)),
  y$samples,
  by = "Run"
)

# Add display labels
pca$study <- stringr::str_to_title(pca$study)
pca$treatment_label <- case_when(
  pca$study == "Huang"                         & pca$treatment == "HP"      ~ "Clade 2.3.4.4",
  pca$study %in% c("Smith", "Campbell")        & pca$treatment == "HP"      ~ "Clade 1",
  pca$study == "Morris"                        & pca$treatment == "HP_2.2"  ~ "Clade 2.2.2.1",
  pca$study == "Morris"                        & pca$treatment == "HP_2.3.2.1" ~ "Clade 2.3.2.1a",
  pca$treatment == "control"                                                 ~ "Control",
  TRUE                                                                       ~ pca$treatment
)

# PCA coloured by study, shaped by treatment
p_pca1 <- ggplot(pca, aes(PC1, PC2, color = study, shape = treatment_label)) +
  geom_point(size = 3) +
  stat_ellipse() +
  labs(x = "PC1", y = "PC2", shape = "Treatment", color = "Study") +
  theme_minimal()

print(p_pca1)
ggsave("supplementary_figure_1.png", p_pca1, width = 12, height = 8, dpi = 300)

# PCA coloured by DPI, faceted by treatment
ggplot(pca, aes(PC1, PC2, color = factor(dpi), shape = treatment_label)) +
  geom_point(size = 3) +
  stat_ellipse() +
  facet_wrap(~ treatment_label) +
  scale_color_viridis_d() +
  labs(x = "PC1", y = "PC2", shape = "Treatment", color = "DPI") +
  theme_minimal()

# Per-study PCA
for (s in unique(pca$study)) {
  pca_s <- na.omit(subset(pca, study == s))
  p <- ggplot(pca_s, aes(PC1, PC2, color = factor(dpi), shape = treatment_label)) +
    geom_point(size = 3) +
    scale_color_viridis_d() +
    labs(title = paste("PCA Plot - Study:", s), x = "PC1", y = "PC2") +
    theme_minimal()
  if (nrow(pca_s) >= 3) p <- p + stat_ellipse()
  print(p)
}

rm(pca_res, pca, p, pca_s, s, p_pca1)
```

## 3.2 Sum technical replicates (Morris) and verify

```{r, message=FALSE}
cat("Morris dimensions before summing tech reps:", dim(morris), "\n")
morris <- sumTechReps(morris, ID = morris$samples$sample_id)
cat("Morris dimensions after summing tech reps:",  dim(morris), "\n")

# Quick PCA to verify
log_cpm <- cpm(morris, log = TRUE)
log_cpm <- log_cpm[apply(log_cpm, 1, var) > 0, ]
pca_morris <- data.frame(prcomp(t(log_cpm), scale. = TRUE)$x[, 1:2], morris$samples)

ggplot(pca_morris, aes(PC1, PC2, color = factor(dpi), shape = treatment)) +
  geom_point(size = 3) +
  stat_ellipse(level = 0.95) +
  scale_color_viridis_d() +
  labs(title = "PCA Plot of Morris Dataset (post tech-rep sum)", x = "PC1", y = "PC2") +
  theme_minimal()

rm(log_cpm, pca_morris)
```

## 3.3 Filtering low-count genes

```{r}
filter_and_report <- function(dge) {
  cat("Before filtering:", dim(dge), "\n")
  keep <- filterByExpr(dge, group = dge$samples$group)
  dge  <- dge[keep, , keep.lib.sizes = FALSE]
  cat("After filtering: ", dim(dge), "\n")
  dge
}

cat("campbell\n"); campbell <- filter_and_report(campbell)
cat("huang\n");    huang    <- filter_and_report(huang)
cat("morris\n");   morris   <- filter_and_report(morris)
cat("smith\n");    smith    <- filter_and_report(smith)

rm(filter_and_report)
```

## 3.4 Normalising

```{r}
campbell <- normLibSizes(campbell)
huang    <- normLibSizes(huang)
morris   <- normLibSizes(morris)
smith    <- normLibSizes(smith)
```

## 3.5 PCA after normalisation

```{r, message=FALSE, warning=FALSE}
plotPCA <- function(dge, title) {
  log_cpm <- cpm(dge, log = TRUE)
  log_cpm <- log_cpm[apply(log_cpm, 1, var) > 0, ]
  pca <- data.frame(prcomp(t(log_cpm), scale. = TRUE)$x[, 1:2], dge$samples)
  ggplot(pca, aes(PC1, PC2, color = factor(dpi), shape = factor(treatment))) +
    geom_point(size = 3) +
    stat_ellipse(aes(group = interaction(dpi, treatment)), level = 0.95, linetype = "dashed") +
    ggtitle(title) +
    labs(color = "dpi", shape = "treatment") +
    theme_minimal()
}

print(plotPCA(campbell, "Campbell: PCA Plot"))
print(plotPCA(huang,    "Huang: PCA Plot"))
print(plotPCA(morris,   "Morris: PCA Plot"))
print(plotPCA(smith,    "Smith: PCA Plot"))

rm(y, counts)
```

# 4. Differential expression analysis
## 4.1 Estimate dispersion and fit model

```{r}
make_design <- function(dge) {
  d <- model.matrix(~ 0 + group, data = dge$samples)
  colnames(d) <- gsub("group", "", colnames(d))
  d
}

design_campbell <- make_design(campbell)
campbell <- estimateDisp(campbell, design_campbell)
fit_campbell <- glmFit(campbell, design_campbell)

design_smith <- make_design(smith)
smith <- estimateDisp(smith, design_smith)
fit_smith <- glmFit(smith, design_smith)

design_morris <- make_design(morris)
morris <- estimateDisp(morris, design_morris)
fit_morris <- glmFit(morris, design_morris)

design_huang <- make_design(huang)
# Huang has only one condition so we borrow BCV from the other studies
bcv <- mean(morris$common.dispersion, campbell$common.dispersion, smith$common.dispersion)
fit_huang <- glmFit(huang, design_huang, dispersion = bcv^2)

rm(make_design, bcv)
```

## 4.2 Differential expression

```{r}
contrasts_campbell <- makeContrasts(
  HPvsControl_1dpi = HP_1 - control_1,
  HPvsControl_2dpi = HP_2 - control_2,
  HPvsControl_3dpi = HP_3 - control_3,
  levels = design_campbell
)
campbell_1 <- topTags(glmLRT(fit_campbell, contrast = contrasts_campbell[, "HPvsControl_1dpi"]), n = Inf)$table
campbell_2 <- topTags(glmLRT(fit_campbell, contrast = contrasts_campbell[, "HPvsControl_2dpi"]), n = Inf)$table
campbell_3 <- topTags(glmLRT(fit_campbell, contrast = contrasts_campbell[, "HPvsControl_3dpi"]), n = Inf)$table

contrasts_huang <- makeContrasts(
  HPvsControl_1dpi = HP_1 - control_3,
  HPvsControl_2dpi = HP_2 - control_3,
  HPvsControl_3dpi = HP_3 - control_3,
  levels = design_huang
)
huang_1 <- topTags(glmLRT(fit_huang, contrast = contrasts_huang[, "HPvsControl_1dpi"]), n = Inf)$table
huang_2 <- topTags(glmLRT(fit_huang, contrast = contrasts_huang[, "HPvsControl_2dpi"]), n = Inf)$table
huang_3 <- topTags(glmLRT(fit_huang, contrast = contrasts_huang[, "HPvsControl_3dpi"]), n = Inf)$table

contrasts_smith <- makeContrasts(
  HPvsControl_1dpi = HP_1 - control_1,
  HPvsControl_3dpi = HP_3 - control_3,
  levels = design_smith
)
smith_1 <- topTags(glmLRT(fit_smith, contrast = contrasts_smith[, "HPvsControl_1dpi"]), n = Inf)$table
smith_3 <- topTags(glmLRT(fit_smith, contrast = contrasts_smith[, "HPvsControl_3dpi"]), n = Inf)$table

contrasts_morris <- makeContrasts(
  HPvsControl_2.2_2dpi   = HP_2.2_2   - control_0,
  HPvsControl_2.3.2.1_2dpi = HP_2.3.2.1_2 - control_0,
  levels = design_morris
)
morris_2.2_2     <- topTags(glmLRT(fit_morris, contrast = contrasts_morris[, "HPvsControl_2.2_2dpi"]),     n = Inf)$table
morris_2.3.2.1_2 <- topTags(glmLRT(fit_morris, contrast = contrasts_morris[, "HPvsControl_2.3.2.1_2dpi"]), n = Inf)$table
```

## 4.3 Venn diagrams

```{r}
# Helper: gene IDs passing a criterion
sig_genes <- function(df, col, threshold = 0.05, direction = NULL) {
  mask <- df[[col]] <= threshold
  if (!is.null(direction)) mask <- mask & (if (direction == "up") df$logFC > 0 else df$logFC < 0)
  rownames(df)[mask]
}

# P-value Venns
ggVennDiagram(list(
  `Morris 2.2 2dpi`   = sig_genes(morris_2.2_2,     "PValue"),
  `Morris 2.3.2.1 2dpi` = sig_genes(morris_2.3.2.1_2, "PValue"),
  `Campbell 2dpi`     = sig_genes(campbell_2,        "PValue"),
  `Huang 2dpi`        = sig_genes(huang_2,           "PValue")
)) + scale_fill_gradient(low = "white", high = "blue")

ggVennDiagram(list(
  `Campbell 1dpi` = sig_genes(campbell_1, "PValue"),
  `Huang 1dpi`    = sig_genes(huang_1,    "PValue"),
  `Smith 1dpi`    = sig_genes(smith_1,    "PValue")
)) + scale_fill_gradient(low = "white", high = "red")

ggVennDiagram(list(
  `Campbell 3dpi` = sig_genes(campbell_3, "PValue"),
  `Huang 3dpi`    = sig_genes(huang_3,    "PValue"),
  `Smith 3dpi`    = sig_genes(smith_3,    "PValue")
)) + scale_fill_gradient(low = "white", high = "purple")

# FDR Venns
ggVennDiagram(list(
  `Morris 2.2 2dpi`     = sig_genes(morris_2.2_2,     "FDR"),
  `Morris 2.3.2.1 2dpi` = sig_genes(morris_2.3.2.1_2, "FDR"),
  `Campbell 2dpi`       = sig_genes(campbell_2,        "FDR"),
  `Huang 2dpi`          = sig_genes(huang_2,           "FDR")
)) + scale_fill_gradient(low = "white", high = "blue")

ggVennDiagram(list(
  `Campbell 1dpi` = sig_genes(campbell_1, "FDR"),
  `Huang 1dpi`    = sig_genes(huang_1,    "FDR"),
  `Smith 1dpi`    = sig_genes(smith_1,    "FDR")
)) + scale_fill_gradient(low = "white", high = "red")

ggVennDiagram(list(
  `Campbell 3dpi` = sig_genes(campbell_3, "FDR"),
  `Huang 3dpi`    = sig_genes(huang_3,    "FDR"),
  `Smith 3dpi`    = sig_genes(smith_3,    "FDR")
)) + scale_fill_gradient(low = "white", high = "purple")

# FDR up-regulated Venns
ggVennDiagram(list(
  `Morris 2.2 2dpi`     = sig_genes(morris_2.2_2,     "FDR", direction = "up"),
  `Morris 2.3.2.1 2dpi` = sig_genes(morris_2.3.2.1_2, "FDR", direction = "up"),
  `Campbell 2dpi`       = sig_genes(campbell_2,        "FDR", direction = "up"),
  `Huang 2dpi`          = sig_genes(huang_2,           "FDR", direction = "up")
)) + scale_fill_gradient(low = "white", high = "blue")

ggVennDiagram(list(
  `Campbell 1dpi` = sig_genes(campbell_1, "FDR", direction = "up"),
  `Huang 1dpi`    = sig_genes(huang_1,    "FDR", direction = "up"),
  `Smith 1dpi`    = sig_genes(smith_1,    "FDR", direction = "up")
)) + scale_fill_gradient(low = "white", high = "red")

ggVennDiagram(list(
  `Campbell 3dpi` = sig_genes(campbell_3, "FDR", direction = "up"),
  `Huang 3dpi`    = sig_genes(huang_3,    "FDR", direction = "up"),
  `Smith 3dpi`    = sig_genes(smith_3,    "FDR", direction = "up")
)) + scale_fill_gradient(low = "white", high = "purple")

# FDR down-regulated Venns
ggVennDiagram(list(
  `Morris 2.2 2dpi`     = sig_genes(morris_2.2_2,     "FDR", direction = "down"),
  `Morris 2.3.2.1 2dpi` = sig_genes(morris_2.3.2.1_2, "FDR", direction = "down"),
  `Campbell 2dpi`       = sig_genes(campbell_2,        "FDR", direction = "down"),
  `Huang 2dpi`          = sig_genes(huang_2,           "FDR", direction = "down")
)) + scale_fill_gradient(low = "white", high = "blue")

ggVennDiagram(list(
  `Campbell 1dpi` = sig_genes(campbell_1, "FDR", direction = "down"),
  `Huang 1dpi`    = sig_genes(huang_1,    "FDR", direction = "down"),
  `Smith 1dpi`    = sig_genes(smith_1,    "FDR", direction = "down")
)) + scale_fill_gradient(low = "white", high = "red")

ggVennDiagram(list(
  `Campbell 3dpi` = sig_genes(campbell_3, "FDR", direction = "down"),
  `Huang 3dpi`    = sig_genes(huang_3,    "FDR", direction = "down"),
  `Smith 3dpi`    = sig_genes(smith_3,    "FDR", direction = "down")
)) + scale_fill_gradient(low = "white", high = "purple")

rm(sig_genes)
```

# 5. P-value combination and jackknife sensitivity analysis
## 5.1 P-value distributions under the null

```{r}
for (df_name in c("smith_1", "smith_3",
                  "campbell_1", "campbell_2", "campbell_3",
                  "morris_2.3.2.1_2", "morris_2.2_2",
                  "huang_1", "huang_2", "huang_3")) {
  hist(get(df_name)$PValue, 20, main = df_name, xlab = "P-value")
}
```

## 5.2 Meta-analysis function and p-value combination

```{r}
metaAnalysis <- function(study_list, study_names, weights) {
  common_genes <- Reduce(intersect, lapply(study_list, rownames))
  gene_name_map <- setNames(annotation_data$external_gene_name, annotation_data$ensembl_gene_id)

  out <- data.frame(gene = common_genes,
                    p_combined = NA_real_,
                    eff_direction = NA_character_,
                    same_direction = NA,
                    studies_included = NA_character_,
                    stringsAsFactors = FALSE)
  for (s in study_names) {
    out[[paste0("logFC_",  s)]] <- NA_real_
    out[[paste0("PValue_", s)]] <- NA_real_
    out[[paste0("FDR_",    s)]] <- NA_real_
  }

  for (i in seq_along(common_genes)) {
    gene <- common_genes[i]
    p_values <- fc_values <- numeric()
    studies  <- character()

    for (j in seq_along(study_list)) {
      df    <- study_list[[j]]
      study <- study_names[j]
      out[i, paste0("logFC_",  study)] <- df[gene, "logFC"]
      out[i, paste0("PValue_", study)] <- df[gene, "PValue"]
      out[i, paste0("FDR_",    study)] <- df[gene, "FDR"]
      p_values  <- c(p_values,  df[gene, "PValue"])
      fc_values <- c(fc_values, df[gene, "logFC"])
      studies   <- c(studies,   study)
    }

    wf <- wFisher(p = p_values, weight = weights,
                  is.onetail = FALSE, eff.sign = sign(fc_values))
    out$p_combined[i]     <- wf$p
    out$eff_direction[i]  <- wf$overall.eff.direction
    out$same_direction[i] <- (all(fc_values > 0) || all(fc_values < 0))
    out$studies_included[i] <- paste(studies, collapse = ", ")
  }

  out$gene_name    <- gene_name_map[out$gene]
  out$FDR_combined <- p.adjust(out$p_combined, method = "BH")
  col_order <- c("gene", "gene_name", "p_combined", "FDR_combined",
                 "same_direction", "studies_included",
                 setdiff(names(out), c("gene", "gene_name", "p_combined",
                                       "FDR_combined", "same_direction",
                                       "studies_included")))
  out[order(out$FDR_combined), col_order]
}

combined_1 <- metaAnalysis(
  study_list   = list(campbell_1, smith_1, huang_1),
  study_names  = c("campbell_1", "smith_1", "huang_1"),
  weights      = c(4, 3, 1)
)
combined_2 <- metaAnalysis(
  study_list   = list(morris_2.2_2, morris_2.3.2.1_2, campbell_2, huang_2),
  study_names  = c("morris_2.2_2", "morris_2.3.2.1_2", "campbell_2", "huang_2"),
  weights      = c(3, 3, 4, 1)
)
combined_3 <- metaAnalysis(
  study_list   = list(campbell_3, smith_3, huang_3),
  study_names  = c("campbell_3", "smith_3", "huang_3"),
  weights      = c(4, 3, 1)
)
print(head(combined_3))
```

## 5.3 Jackknife sampling

```{r}
jackknife_meta2 <- function(study_list, study_names, weights) {
  setNames(
    lapply(seq_along(study_list), function(i) {
      metaAnalysis(study_list[-i], study_names[-i], weights[-i])
    }),
    paste0("jackknife_excluding_", study_names)
  )
}

jackknife_results_1 <- jackknife_meta2(
  study_list  = list(campbell_1 = campbell_1, smith_1 = smith_1, huang_1 = huang_1),
  study_names = c("campbell_1", "smith_1", "huang_1"),
  weights     = c(4, 3, 1)
)

jackknife_results_2 <- jackknife_meta2(
  study_list  = list(morris_2.2_2 = morris_2.2_2, morris_2.3.2.1_2 = morris_2.3.2.1_2,
                     campbell_2 = campbell_2, huang_2 = huang_2),
  study_names = c("morris_2.2_2", "morris_2.3.2.1_2", "campbell_2", "huang_2"),
  weights     = c(3, 3, 4, 1)
)

jackknife_results_3 <- jackknife_meta2(
  study_list  = list(campbell_3 = campbell_3, smith_3 = smith_3, huang_3 = huang_3),
  study_names = c("campbell_3", "smith_3", "huang_3"),
  weights     = c(4, 3, 1)
)
```

## 5.4 Build jackknife flags

```{r}
# Helper to join jackknife results and compute flags
add_jackknife_flags_1_3 <- function(combined, jack_results, day,
                                    study_cols, day_suffix) {
  # Study-specific FDR column names in combined
  fdr_cols <- paste0("FDR_", study_cols, "_", day_suffix)

  combined %>%
    left_join(
      jack_results[[paste0("jackknife_excluding_campbell_", day)]] %>%
        dplyr::select(gene,
               p_combined_no_campbell   = p_combined,
               FDR_combined_no_campbell = FDR_combined,
               same_direction_no_campbell = same_direction,
               eff_direction_no_campbell  = eff_direction),
      by = "gene"
    ) %>%
    left_join(
      jack_results[[paste0("jackknife_excluding_smith_", day)]] %>%
        dplyr::select(gene,
               p_combined_no_smith   = p_combined,
               FDR_combined_no_smith = FDR_combined,
               same_direction_no_smith = same_direction,
               eff_direction_no_smith  = eff_direction),
      by = "gene"
    ) %>%
    left_join(
      jack_results[[paste0("jackknife_excluding_huang_", day)]] %>%
        dplyr::select(gene,
               p_combined_no_huang   = p_combined,
               FDR_combined_no_huang = FDR_combined,
               same_direction_no_huang = same_direction,
               eff_direction_no_huang  = eff_direction),
      by = "gene"
    ) %>%
    mutate(
      jackknife = as.integer(
        eff_direction_no_campbell == eff_direction_no_smith &
        eff_direction_no_campbell == eff_direction_no_huang &
        FDR_combined_no_campbell < 0.05 &
        FDR_combined_no_smith    < 0.05 &
        FDR_combined_no_huang    < 0.05
      ),
      combined = as.integer(FDR_combined < 0.05),
      campbell = as.integer(.data[[paste0("FDR_campbell_", day_suffix)]] < 0.05),
      smith    = as.integer(.data[[paste0("FDR_smith_",    day_suffix)]] < 0.05),
      huang    = as.integer(.data[[paste0("FDR_huang_",    day_suffix)]] < 0.05)
    )
}

combined_1_jackknife <- add_jackknife_flags_1_3(combined_1, jackknife_results_1,
                                                 day = 1, day_suffix = 1,
                                                 study_cols = c("campbell", "smith", "huang"))
combined_3_jackknife <- add_jackknife_flags_1_3(combined_3, jackknife_results_3,
                                                 day = 3, day_suffix = 3,
                                                 study_cols = c("campbell", "smith", "huang"))

# Day 2 has 4 studies (two Morris sub-strains)
combined_2_jackknife <- combined_2 %>%
  left_join(
    jackknife_results_2$jackknife_excluding_campbell_2 %>%
      dplyr::select(gene,
             p_combined_no_campbell   = p_combined,
             FDR_combined_no_campbell = FDR_combined,
             same_direction_no_campbell = same_direction,
             eff_direction_no_campbell  = eff_direction),
    by = "gene"
  ) %>%
  left_join(
    jackknife_results_2$jackknife_excluding_huang_2 %>%
      dplyr::select(gene,
             p_combined_no_huang   = p_combined,
             FDR_combined_no_huang = FDR_combined,
             same_direction_no_huang = same_direction,
             eff_direction_no_huang  = eff_direction),
    by = "gene"
  ) %>%
  left_join(
    jackknife_results_2$`jackknife_excluding_morris_2.3.2.1_2` %>%
      dplyr::select(gene,
             p_combined_no_morris_2_3_2_1   = p_combined,
             FDR_combined_no_morris_2_3_2_1 = FDR_combined,
             same_direction_no_morris_2_3_2_1 = same_direction,
             eff_direction_no_morris_2_3_2_1  = eff_direction),
    by = "gene"
  ) %>%
  left_join(
    jackknife_results_2$`jackknife_excluding_morris_2.2_2` %>%
      dplyr::select(gene,
             p_combined_no_morris_2_2   = p_combined,
             FDR_combined_no_morris_2_2 = FDR_combined,
             same_direction_no_morris_2_2 = same_direction,
             eff_direction_no_morris_2_2  = eff_direction),
    by = "gene"
  ) %>%
  mutate(
    jackknife = as.integer(
      eff_direction_no_campbell == eff_direction_no_huang &
      eff_direction_no_campbell == eff_direction_no_morris_2_3_2_1 &
      eff_direction_no_campbell == eff_direction_no_morris_2_2 &
      FDR_combined_no_campbell      < 0.05 &
      FDR_combined_no_huang         < 0.05 &
      FDR_combined_no_morris_2_3_2_1 < 0.05 &
      FDR_combined_no_morris_2_2    < 0.05
    ),
    combined       = as.integer(FDR_combined < 0.05),
    campbell       = as.integer(FDR_campbell_2     < 0.05),
    huang          = as.integer(FDR_huang_2         < 0.05),
    morris_2_3_2_1 = as.integer(`FDR_morris_2.3.2.1_2` < 0.05),
    morris_2_2     = as.integer(`FDR_morris_2.2_2`     < 0.05)
  )

rm(add_jackknife_flags_1_3)
```

## 5.5 UpSetR plots

```{r}
upset(combined_1_jackknife, sets = c("smith",    "campbell", "huang",    "combined"))
upset(combined_2_jackknife, sets = c("huang",    "campbell", "morris_2_2", "morris_2_3_2_1", "combined"))
upset(combined_3_jackknife, sets = c("smith",    "campbell", "huang",    "combined"))

upset(combined_1_jackknife, sets = c("jackknife", "smith",  "campbell", "huang"))
upset(combined_2_jackknife, sets = c("jackknife", "huang",  "campbell", "morris_2_2", "morris_2_3_2_1"))
upset(combined_3_jackknife, sets = c("jackknife", "smith",  "campbell", "huang"))
```

## 5.6 UpSetR for day 3 with GO annotations

```{r}
annotation_data_renamed <- read.csv("Anas_platyrhynchos2_annotation.csv")
immune     <- GOBPOFFSPRING[["GO:0006955"]]
cell_death <- GOBPOFFSPRING[["GO:0008219"]]

# Rename columns for display
combined_3_renamed <- combined_3_jackknife
colnames(combined_3_renamed)[colnames(combined_3_renamed) == "campbell"]  <- "Campbell"
colnames(combined_3_renamed)[colnames(combined_3_renamed) == "jackknife"] <- "Jackknife"
colnames(combined_3_renamed)[colnames(combined_3_renamed) == "smith"]     <- "Smith"
colnames(combined_3_renamed)[colnames(combined_3_renamed) == "huang"]     <- "Huang"

annotation_collapsed <- annotation_data_renamed %>%
  group_by(ensembl_gene_id) %>%
  summarise(GO_terms = str_c(unique(go_id), collapse = ","), .groups = "drop")

combined_3_renamed2 <- combined_3_renamed %>%
  left_join(annotation_collapsed, by = c("gene" = "ensembl_gene_id")) %>%
  mutate(
    GO_terms  = replace_na(GO_terms, ""),
    GO_list   = str_split(GO_terms, ","),
    has_immune     = map_lgl(GO_list, ~ any(.x %in% immune)),
    has_cell_death = map_lgl(GO_list, ~ any(.x %in% cell_death)),
    category = case_when(
      has_immune & has_cell_death ~ "both",
      has_immune                  ~ "immune",
      has_cell_death              ~ "cell_death",
      TRUE                        ~ NA_character_
    )
  ) %>%
  dplyr::select(-has_immune, -has_cell_death, -GO_list)

CategoryQuery <- function(row, target_category) row["category"] == target_category

upset(combined_3_renamed, sets = c("Jackknife", "Campbell", "Smith", "Huang"),
      mainbar.y.label = "Number of differentially expressed genes",
      sets.x.label    = "Differentially expressed genes per study")

upset(combined_3_renamed2, sets = c("Jackknife", "Campbell", "Smith", "Huang"),
      mainbar.y.label = "Number of differentially expressed genes",
      sets.x.label    = "Differentially expressed genes per study")

upset(combined_3_renamed2,
      sets = c("Jackknife", "Campbell", "Smith", "Huang"),
      mainbar.y.label = "Number of differentially expressed genes",
      sets.x.label    = "Differentially expressed genes per study",
      queries = list(
        list(query = CategoryQuery, params = list("immune"),
             color = "blue",  active = TRUE, query.name = "Immune genes"),
        list(query = CategoryQuery, params = list("cell_death"),
             color = "green", active = TRUE, query.name = "Death genes"),
        list(query = CategoryQuery, params = list("both"),
             color = "pink",  active = TRUE, query.name = "Both")
      ),
      query.legend = "bottom")

# High-resolution export for day 3
tiff("upset.tiff", width = 10, height = 6, units = "in", res = 1000, compression = "lzw")
upset(
  combined_3_renamed2,
  sets            = c("Jackknife", "Smith", "Campbell", "Huang"),
  keep.order      = TRUE,
  mainbar.y.label = "Number of differentially expressed genes per intersection",
  sets.x.label    = "Total DEGs per study",
  queries = list(
    list(query = intersects, params = list("Jackknife"),
         color = "#c95117", active = TRUE),
    list(query = intersects, params = list("Jackknife", "Campbell", "Smith", "Huang"),
         color = "#f29d00", active = TRUE)
  )
)
dev.off()
```

## 5.7 UpSetR with combined flag (saved PNGs)

```{r}
orig1 <- colnames(combined_1_jackknife)
orig2 <- colnames(combined_2_jackknife)
orig3 <- colnames(combined_3_jackknife)

rename_cols <- function(df, old, new) {
  colnames(df)[colnames(df) %in% old] <- new
  df
}

combined_1_jackknife <- rename_cols(combined_1_jackknife,
  c("jackknife","smith","campbell","huang","combined"),
  c("Jackknife","Smith","Campbell","Huang","Combined"))

combined_2_jackknife <- rename_cols(combined_2_jackknife,
  c("jackknife","huang","campbell","morris_2_2","morris_2_3_2_1","combined"),
  c("Jackknife","Huang","Campbell","Morris_2_2","Morris_2_3_2_1","Combined"))

combined_3_jackknife <- rename_cols(combined_3_jackknife,
  c("jackknife","smith","campbell","huang","combined"),
  c("Jackknife","Smith","Campbell","Huang","Combined"))

png("upset_combined_1.png", width = 1200, height = 800, res = 150)
upset(combined_1_jackknife, sets = c("Jackknife","Smith","Campbell","Huang","Combined"))
dev.off()

png("upset_combined_2.png", width = 1200, height = 800, res = 150)
upset(combined_2_jackknife, sets = c("Jackknife","Huang","Campbell","Morris_2_2","Morris_2_3_2_1","Combined"))
dev.off()

png("upset_combined_3.png", width = 1200, height = 800, res = 150)
upset(combined_3_jackknife, sets = c("Jackknife","Smith","Campbell","Huang","Combined"))
dev.off()

# Restore original column names
colnames(combined_1_jackknife) <- orig1
colnames(combined_2_jackknife) <- orig2
colnames(combined_3_jackknife) <- orig3
rm(orig1, orig2, orig3, rename_cols)
```

# 6. Interactive datatables

```{r, message=FALSE}
datatable(combined_1_jackknife)
datatable(combined_2_jackknife)
datatable(combined_3_jackknife)
```

```{r}
datatable(combined_1_jackknife %>% filter(combined == 1, smith != 1, huang != 1, campbell != 1))
datatable(combined_2_jackknife %>% filter(combined == 1, morris_2_2 != 1, morris_2_3_2_1 != 1, huang != 1, campbell != 1))
datatable(combined_3_jackknife %>% filter(combined == 1, smith != 1, huang != 1, campbell != 1))
```

```{r}
datatable(combined_1_jackknife %>% filter(combined == 1))
datatable(combined_2_jackknife %>% filter(combined == 1))
datatable(combined_3_jackknife %>% filter(combined == 1))
```

```{r}
datatable(combined_1_jackknife %>% filter(combined == 1, smith == 1, huang == 1, campbell == 1))
datatable(combined_2_jackknife %>% filter(combined == 1, morris_2_2 == 1, morris_2_3_2_1 == 1, huang == 1, campbell == 1))
datatable(combined_3_jackknife %>% filter(combined == 1, smith == 1, huang == 1, campbell == 1))
```

```{r}
datatable(combined_1_jackknife %>% filter(jackknife == 1, smith != 1, huang != 1, campbell != 1))
datatable(combined_2_jackknife %>% filter(jackknife == 1, morris_2_2 != 1, morris_2_3_2_1 != 1, huang != 1, campbell != 1))
datatable(combined_3_jackknife %>% filter(jackknife == 1, smith != 1, huang != 1, campbell != 1))
```

```{r}
datatable(combined_1_jackknife %>% filter(jackknife == 1))
datatable(combined_2_jackknife %>% filter(jackknife == 1))
datatable(combined_3_jackknife %>% filter(jackknife == 1))
```

```{r}
datatable(combined_1_jackknife %>% filter(jackknife == 1, smith == 1, huang == 1, campbell == 1))
datatable(combined_2_jackknife %>% filter(jackknife == 1, morris_2_2 == 1, morris_2_3_2_1 == 1, huang == 1, campbell == 1))
datatable(combined_3_jackknife %>% filter(jackknife == 1, smith == 1, huang == 1, campbell == 1))
```

# 7. Log fold-change bar plots
## 7.1 Helper function and shared theme

```{r}
# Shared bar-plot helper to avoid repeating pivot+ggplot code
barplot_logfc <- function(df, gene_col, logfc_cols, title,
                          filter_expr = TRUE, n = 20) {
  df %>%
    filter(!is.na(gene_name) & gene_name != "", {{ filter_expr }}) %>%
    slice_head(n = n) %>%
    dplyr::select(all_of(c(gene_col, logfc_cols))) %>%
    pivot_longer(-all_of(gene_col), names_to = "source", values_to = "logFC") %>%
    ggplot(aes(x = .data[[gene_col]], y = logFC, fill = source)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title = title, x = "Gene", y = "logFC") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## 7.2 Top 20 combined genes

```{r}
barplot_logfc(combined_1_jackknife, "gene_name",
              c("logFC_campbell_1","logFC_smith_1","logFC_huang_1"),
              "Log Fold Changes for 1dpi",
              filter_expr = combined == 1)

barplot_logfc(combined_2_jackknife, "gene_name",
              c("logFC_morris_2.2_2","logFC_morris_2.3.2.1_2","logFC_campbell_2","logFC_huang_2"),
              "Log Fold Changes for 2dpi",
              filter_expr = combined == 1)

barplot_logfc(combined_3_jackknife, "gene_name",
              c("logFC_campbell_3","logFC_smith_3","logFC_huang_3"),
              "Log Fold Changes for 3dpi",
              filter_expr = combined == 1)
```

## 7.3 Top 20 common combined genes (in all studies)

```{r}
barplot_logfc(combined_1_jackknife, "gene_name",
              c("logFC_campbell_1","logFC_smith_1","logFC_huang_1"),
              "Log Fold Changes for 1dpi (all studies)",
              filter_expr = combined == 1 & campbell == 1 & smith == 1 & huang == 1)

barplot_logfc(combined_2_jackknife, "gene_name",
              c("logFC_morris_2.2_2","logFC_morris_2.3.2.1_2","logFC_campbell_2","logFC_huang_2"),
              "Log Fold Changes for 2dpi (all studies)",
              filter_expr = combined == 1 & campbell == 1 & morris_2_3_2_1 == 1 & morris_2_2 == 1 & huang == 1)

barplot_logfc(combined_3_jackknife, "gene_name",
              c("logFC_campbell_3","logFC_smith_3","logFC_huang_3"),
              "Log Fold Changes for 3dpi (all studies)",
              filter_expr = combined == 1 & campbell == 1 & smith == 1 & huang == 1)
```

## 7.4 Top 20 unique combined genes

```{r}
barplot_logfc(combined_1_jackknife, "gene_name",
              c("logFC_campbell_1","logFC_smith_1","logFC_huang_1"),
              "Log Fold Changes for 1dpi (unique to combined)",
              filter_expr = combined == 1 & campbell != 1 & smith != 1 & huang != 1)

barplot_logfc(combined_2_jackknife, "gene_name",
              c("logFC_morris_2.2_2","logFC_morris_2.3.2.1_2","logFC_campbell_2","logFC_huang_2"),
              "Log Fold Changes for 2dpi (unique to combined)",
              filter_expr = combined == 1 & campbell != 1 & morris_2_3_2_1 != 1 & morris_2_2 != 1 & huang != 1)

barplot_logfc(combined_3_jackknife, "gene_name",
              c("logFC_campbell_3","logFC_smith_3","logFC_huang_3"),
              "Log Fold Changes for 3dpi (unique to combined)",
              filter_expr = combined == 1 & campbell != 1 & smith != 1 & huang != 1)
```

## 7.5 Top 20 jackknife genes

```{r}
barplot_logfc(combined_1_jackknife, "gene_name",
              c("logFC_campbell_1","logFC_smith_1","logFC_huang_1"),
              "Log Fold Changes for 1dpi (jackknife)",
              filter_expr = jackknife == 1)

barplot_logfc(combined_2_jackknife, "gene_name",
              c("logFC_morris_2.2_2","logFC_morris_2.3.2.1_2","logFC_campbell_2","logFC_huang_2"),
              "Log Fold Changes for 2dpi (jackknife)",
              filter_expr = jackknife == 1)

barplot_logfc(combined_3_jackknife, "gene_name",
              c("logFC_campbell_3","logFC_smith_3","logFC_huang_3"),
              "Log Fold Changes for 3dpi (jackknife)",
              filter_expr = jackknife == 1)
```

## 7.6 Top 20 jackknife × common

```{r}
barplot_logfc(combined_1_jackknife, "gene_name",
              c("logFC_campbell_1","logFC_smith_1","logFC_huang_1"),
              "Log Fold Changes for 1dpi (jackknife, all studies)",
              filter_expr = jackknife == 1 & campbell == 1 & smith == 1 & huang == 1)

barplot_logfc(combined_2_jackknife, "gene_name",
              c("logFC_morris_2.2_2","logFC_morris_2.3.2.1_2","logFC_campbell_2","logFC_huang_2"),
              "Log Fold Changes for 2dpi (jackknife, all studies)",
              filter_expr = jackknife == 1 & campbell == 1 & morris_2_3_2_1 == 1 & morris_2_2 == 1 & huang == 1)

barplot_logfc(combined_3_jackknife, "gene_name",
              c("logFC_campbell_3","logFC_smith_3","logFC_huang_3"),
              "Log Fold Changes for 3dpi (jackknife, all studies)",
              filter_expr = jackknife == 1 & campbell == 1 & smith == 1 & huang == 1)
```

## 7.7 Top 20 jackknife unique genes

```{r}
barplot_logfc(combined_1_jackknife, "gene_name",
              c("logFC_campbell_1","logFC_smith_1","logFC_huang_1"),
              "Log Fold Changes for 1dpi (jackknife unique)",
              filter_expr = jackknife == 1 & campbell != 1 & smith != 1 & huang != 1)

barplot_logfc(combined_2_jackknife, "gene_name",
              c("logFC_morris_2.2_2","logFC_morris_2.3.2.1_2","logFC_campbell_2","logFC_huang_2"),
              "Log Fold Changes for 2dpi (jackknife unique)",
              filter_expr = jackknife == 1 & campbell != 1 & morris_2_3_2_1 != 1 & morris_2_2 != 1 & huang != 1)

barplot_logfc(combined_3_jackknife, "gene_name",
              c("logFC_campbell_3","logFC_smith_3","logFC_huang_3"),
              "Log Fold Changes for 3dpi (jackknife unique)",
              filter_expr = jackknife == 1 & campbell != 1 & smith != 1 & huang != 1)

rm(barplot_logfc)
```

## 7.8 Publication bar plot — selected genes, day 3

```{r}
label_colors <- c(
  "Jackknife_only" = "#C95117",
  "All_four"       = "#f29d00",
  "Other"          = "black"
)

plot_df <- combined_3_renamed2 %>%
  mutate(
    gene_name = as.character(gene_name),
    Jackknife = as.integer(unlist(Jackknife)),
    Smith     = as.integer(unlist(Smith)),
    Huang     = as.integer(unlist(Huang)),
    Campbell  = as.integer(unlist(Campbell)),
    label_group = case_when(
      Jackknife == 1 & Smith == 0 & Huang == 0 & Campbell == 0 ~ "Jackknife_only",
      Jackknife == 1 & Smith == 1 & Huang == 1 & Campbell == 1 ~ "All_four",
      TRUE ~ "Other"
    )
  ) %>%
  filter(!is.na(gene_name) & gene_name != "",
         gene_name %in% c("MX1","OASL","RIGI","STAT4","IFIH1",
                          "RAB27A","TRAF3IP3","ADAR","RSAD2","SRGN",
                          "HMOX1","FBH1","FSTL1","EXT1","FADD"),
         Jackknife == 1) %>%
  dplyr::select(gene_name, label_group, logFC_campbell_3, logFC_smith_3, logFC_huang_3) %>%
  pivot_longer(-c(gene_name, label_group), names_to = "source", values_to = "logFC") %>%
  mutate(
    source     = recode(source,
                        "logFC_campbell_3" = "Campbell",
                        "logFC_smith_3"    = "Smith",
                        "logFC_huang_3"    = "Huang"),
    label_group = as.character(label_group)
  )

dummy_legend_points <- plot_df %>%
  group_by(label_group) %>%
  summarise(gene_name   = dplyr::first(as.character(gene_name)),
            y_for_legend = min(logFC, na.rm = TRUE)) %>%
  ungroup()

ggplot(plot_df, aes(x = gene_name, y = logFC, fill = source)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_point(data = dummy_legend_points,
             aes(x = gene_name, y = y_for_legend, color = label_group),
             inherit.aes = FALSE, size = 4, alpha = 0) +
  scale_fill_manual(name = "Study",
                    values = c("Campbell" = rgb(58, 170, 53, maxColorValue = 255),
                               "Smith"    = rgb(0, 102, 51, maxColorValue = 255),
                               "Huang"    = rgb(188, 207, 0, maxColorValue = 255))) +
  scale_color_manual(name = "Significant",
                     values = label_colors,
                     labels = c("Jackknife_only" = "Exclusively Meta-analysis",
                                "All_four"       = "All Studies & Meta-analysis",
                                "Other"          = "At least one study & Meta-analysis")) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 5, shape = 15), order = 2),
         fill  = guide_legend(order = 1)) +
  labs(title    = "Meta-analysis of duck lung RNA-seq after HPAI H5N1 infection (3 studies)",
       subtitle = "Subset of genes retained after jackknife sensitivity analysis — 3 days post infection",
       x = "Gene",
       y = "Log fold-change in expression after HPAI H5N1 treatment") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45, hjust = 1,
    colour = setNames(c("#C95117", "#f29d00", "black"),
                      c("Jackknife_only", "All_four", "Other"))[
      plot_df %>% distinct(gene_name, label_group) %>%
        arrange(gene_name) %>% pull(label_group)
    ]
  ))

ggsave("barplot.png", device = "tiff", dpi = 1000)
rm(dummy_legend_points, label_colors, plot_df)
```

## 7.9 Virus-related genes across days (GO-filtered bar plots)

```{r}
# Build GO annotations lookup
annotation_data <- read.csv("Anas_platyrhynchos2_annotation.csv")
TERM2GENE <- annotation_data %>%
  dplyr::select(go_id, ensembl_gene_id) %>%
  filter(!is.na(go_id), !is.na(ensembl_gene_id)) %>%
  distinct()
TERM2NAME <- if ("name_1006" %in% colnames(annotation_data)) {
  annotation_data %>% dplyr::select(go_id, name_1006) %>% distinct()
} else NULL

GO_annotations <- TERM2GENE %>%
  left_join(TERM2NAME, by = "go_id") %>%
  group_by(ensembl_gene_id) %>%
  summarise(go_terms = paste(unique(name_1006), collapse = ", "), .groups = "drop")

make_plotdata <- function(tbl, day, n_keep = 100) {
  dyn_cols <- names(tbl)[grepl(paste0("^logFC_.*_", day, "$"), names(tbl))]
  tbl %>%
    dplyr::filter(jackknife == "1", !is.na(gene_name), gene_name != "") %>%
    left_join(GO_annotations, by = c("gene" = "ensembl_gene_id")) %>%
    dplyr::filter(!is.na(go_terms) & grepl("virus", go_terms, ignore.case = TRUE)) %>%
    slice_head(n = n_keep) %>%
    pivot_longer(cols = all_of(dyn_cols), names_to = "study_raw", values_to = "logFC") %>%
    mutate(
      study_key = sub(paste0("_", day, "$"), "", sub("^logFC_", "", study_raw)),
      study_key = gsub("[\\.]", "_", study_key),
      study = case_when(
        grepl("^morris_", study_key) ~ tools::toTitleCase(gsub("_", ".", sub("^morris_", "Morris ", study_key))),
        TRUE ~ tools::toTitleCase(gsub("_", " ", study_key))
      )
    ) %>%
    rowwise() %>%
    mutate(
      significant = if (study_key %in% names(cur_data())) {
        val <- cur_data()[[study_key]]
        !is.na(val) && val != 0
      } else TRUE,
      legend_lab = paste(study, ifelse(significant, "Significant", "Not significant"), sep = " — ")
    ) %>%
    ungroup() %>%
    dplyr::select(gene_name, study, logFC, significant, legend_lab)
}

plotdata1 <- make_plotdata(combined_1_jackknife, day = 1)
plotdata2 <- make_plotdata(combined_2_jackknife, day = 2)
plotdata3 <- make_plotdata(combined_3_jackknife, day = 3)

base_colors <- c(
  "Campbell"       = "#1b9e77",
  "Smith"          = "#d95f02",
  "Huang"          = "#7570b3",
  "Morris 2.2"     = "#e7298a",
  "Morris 2.3.2.1" = "#66a61e"
)

build_fill_values <- function(plotdata, base_colors) {
  studies_present <- unique(plotdata$study)
  cols <- vapply(studies_present, function(s)
    if (s %in% names(base_colors)) base_colors[[s]]
    else scales::hue_pal()(length(studies_present))[which(studies_present == s)],
    character(1))
  unlist(lapply(names(cols), function(s) {
    setNames(c(scales::alpha(cols[[s]], 1), scales::alpha(cols[[s]], 0.3)),
             c(paste0(s, " — Significant"), paste0(s, " — Not significant")))
  }))
}

plot_theme <- theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        plot.title = element_text(face = "bold", hjust = 0.5),
        legend.position = "bottom", legend.direction = "horizontal",
        legend.title = element_text(size = 10), legend.text = element_text(size = 9),
        legend.key.width = unit(1.2, "cm"), legend.key.height = unit(0.5, "cm"))

p_day1 <- ggplot(plotdata1, aes(gene_name, logFC, fill = legend_lab)) +
  geom_col(position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = build_fill_values(plotdata1, base_colors),
                    name = "Study / Statistical significance") +
  labs(x = "Gene", y = "log2 Fold Change",
       title = "Day 1 — log2FC across studies (virus-related genes)") +
  plot_theme + guides(fill = guide_legend(nrow = 3, byrow = TRUE))

p_day2 <- ggplot(plotdata2, aes(gene_name, logFC, fill = legend_lab)) +
  geom_col(position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = build_fill_values(plotdata2, base_colors),
                    name = "Study / Statistical significance") +
  labs(x = "Gene", y = "log2 Fold Change",
       title = "Day 2 — log2FC across studies (virus-related genes)") +
  plot_theme + guides(fill = guide_legend(nrow = 4, byrow = TRUE))

p_day3 <- ggplot(plotdata3, aes(gene_name, logFC, fill = legend_lab)) +
  geom_col(position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = build_fill_values(plotdata3, base_colors),
                    name = "Study / Statistical significance") +
  labs(x = "Gene", y = "log2 Fold Change",
       title = "Day 3 — log2FC across studies (virus-related genes)") +
  plot_theme + guides(fill = guide_legend(nrow = 3, byrow = TRUE))

print(p_day1); ggsave("plot_day1.png", p_day1, width = 16, height = 10, dpi = 300)
print(p_day2); ggsave("plot_day2.png", p_day2, width = 16, height = 10, dpi = 300)
print(p_day3); ggsave("plot_day3.png", p_day3, width = 16, height = 10, dpi = 300)

rm(make_plotdata, build_fill_values, plot_theme, base_colors,
   plotdata1, plotdata2, plotdata3, p_day1, p_day2, p_day3)
```

# 8. Gene Ontology (overrepresentation analysis)
## 8.1 Annotation and helper

```{r}
# TERM2GENE and TERM2NAME already loaded above; reload if needed
if (!exists("TERM2GENE")) {
  annotation_data <- read.csv("Anas_platyrhynchos2_annotation.csv")
  TERM2GENE <- annotation_data %>%
    dplyr::select(go_id, ensembl_gene_id) %>%
    filter(!is.na(go_id), !is.na(ensembl_gene_id)) %>% distinct()
  TERM2NAME <- if ("name_1006" %in% colnames(annotation_data))
    annotation_data %>% dplyr::select(go_id, name_1006) %>% distinct()
  else NA
}

plotgo <- function(target_genes, title = "GO Enrichment") {
  go_enrich <- enricher(gene = target_genes, TERM2GENE = TERM2GENE,
                        TERM2NAME = TERM2NAME, pvalueCutoff = 0.05)
  if (!is.null(go_enrich) && nrow(go_enrich@result) > 0)
    barplot(go_enrich, showCategory = 20, title = title)
  else
    message("No enrichment found.")
}
```

## 8.2 P-value combination genes

```{r, fig.height=15, fig.width=15}
for (day in 1:3) {
  df <- get(paste0("combined_", day, "_jackknife"))
  goi <- df %>% filter(combined == 1) %>% pull(gene) %>% unique() %>% na.omit()
  plotgo(goi, title = sprintf("DEGs after p-value combination, %d dpi", day))
  goi_up   <- df %>% filter(combined == 1, eff_direction == '+') %>% pull(gene) %>% unique() %>% na.omit()
  plotgo(goi_up, title = sprintf("Upregulated after p-value combination, %d dpi", day))
  goi_down <- df %>% filter(combined == 1, eff_direction == '-') %>% pull(gene) %>% unique() %>% na.omit()
  plotgo(goi_down, title = sprintf("Downregulated after p-value combination, %d dpi", day))
}
```

## 8.3 Jackknife genes

```{r, fig.height=15, fig.width=15}
for (day in 1:3) {
  df <- get(paste0("combined_", day, "_jackknife"))
  goi_up   <- df %>% filter(jackknife == 1, eff_direction == '+') %>% pull(gene) %>% unique() %>% na.omit()
  plotgo(goi_up,   title = sprintf("Upregulated after jackknife, %d dpi", day))
  goi_down <- df %>% filter(jackknife == 1, eff_direction == '-') %>% pull(gene) %>% unique() %>% na.omit()
  plotgo(goi_down, title = sprintf("Downregulated after jackknife, %d dpi", day))
}
```

# 9. KEGG enrichment
## 9.1 Gallus gallus ortholog mapping

```{r}
duck_to_gallus   <- read.csv('duck_to_gallus.csv')
annotation_gallus <- read.csv("annotation_gallus_.csv")

process_jack <- function(df, orthology_df, annotation_df) {
  df %>%
    left_join(orthology_df %>% dplyr::select(initial_ensg, ortholog_ensg, ortholog_name),
              by = c("gene" = "initial_ensg")) %>%
    left_join(annotation_df %>% dplyr::select(ensembl_gene_id, entrezgene_id),
              by = c("ortholog_ensg" = "ensembl_gene_id")) %>%
    distinct() %>%
    filter(!is.na(entrezgene_id)) %>%
    arrange(gene, is.na(ortholog_name), str_starts(ortholog_name, "ENSGALG")) %>%
    group_by(gene) %>% dplyr::slice(1) %>% ungroup()
}

combined_1_jackknife_gallus <- process_jack(combined_1_jackknife, duck_to_gallus, annotation_gallus)
combined_2_jackknife_gallus <- process_jack(combined_2_jackknife, duck_to_gallus, annotation_gallus)
combined_3_jackknife_gallus <- process_jack(combined_3_jackknife, duck_to_gallus, annotation_gallus)

plotkegg <- function(target_genes, title, organism) {
  kegg_enrich <- enrichKEGG(gene = target_genes, organism = organism, pvalueCutoff = 0.05)
  if (!is.null(kegg_enrich) && nrow(kegg_enrich@result) > 0)
    barplot(kegg_enrich, showCategory = 20, title = title)
  else
    message("No KEGG enrichment found.")
}
```

## 9.2 KEGG barplots — Gallus gallus

```{r, fig.width=15, fig.height=15}
kegg_subsets <- list(
  list(filter_expr = quote(combined == 1 & eff_direction == '+'), label = "upregulated (combined)"),
  list(filter_expr = quote(combined == 1 & eff_direction == '-'), label = "downregulated (combined)"),
  list(filter_expr = quote(jackknife == 1 & eff_direction == '+'), label = "upregulated (jackknife)"),
  list(filter_expr = quote(jackknife == 1 & eff_direction == '-'), label = "downregulated (jackknife)")
)

for (day in 1:3) {
  df <- get(paste0("combined_", day, "_jackknife_gallus"))
  for (ss in kegg_subsets) {
    goi <- df %>% filter(!!ss$filter_expr) %>% pull(entrezgene_id) %>% unique() %>% na.omit()
    plotkegg(goi, sprintf("%s, %d dpi", ss$label, day), "gga")
  }
}
rm(kegg_subsets)
```

### 9.2.1 Common genes (jackknife, all studies)

```{r, fig.width=15, fig.height=15}
for (day in 1:3) {
  df <- get(paste0("combined_", day, "_jackknife_gallus"))
  filter_expr <- if (day == 2)
    quote(jackknife == 1 & huang == 1 & campbell == 1 & morris_2_3_2_1 == 1 & morris_2_2 == 1)
  else
    quote(jackknife == 1 & huang == 1 & campbell == 1 & smith == 1)
  goi <- df %>% filter(!!filter_expr) %>% pull(entrezgene_id) %>% unique() %>% na.omit()
  plotkegg(goi, sprintf("common genes across all studies (jackknife), %d dpi", day), "gga")
}
```

## 9.3 KEGG barplots — Homo sapiens

```{r}
duck_to_hsapiens  <- read.csv('duck_to_hsapiens.csv')
annotation_hsapiens <- read.csv("annotation_hsapiens_.csv")

combined_1_jackknife_hsapiens <- process_jack(combined_1_jackknife, duck_to_hsapiens, annotation_hsapiens)
combined_2_jackknife_hsapiens <- process_jack(combined_2_jackknife, duck_to_hsapiens, annotation_hsapiens)
combined_3_jackknife_hsapiens <- process_jack(combined_3_jackknife, duck_to_hsapiens, annotation_hsapiens)
```

```{r, fig.width=15, fig.height=15}
kegg_subsets <- list(
  list(filter_expr = quote(combined == 1 & eff_direction == '+'), label = "upregulated (combined)"),
  list(filter_expr = quote(combined == 1 & eff_direction == '-'), label = "downregulated (combined)"),
  list(filter_expr = quote(jackknife == 1 & eff_direction == '+'), label = "upregulated (jackknife)"),
  list(filter_expr = quote(jackknife == 1 & eff_direction == '-'), label = "downregulated (jackknife)")
)

for (day in 1:3) {
  df <- get(paste0("combined_", day, "_jackknife_hsapiens"))
  for (ss in kegg_subsets) {
    goi <- df %>% filter(!!ss$filter_expr) %>% pull(entrezgene_id) %>% unique() %>% na.omit()
    plotkegg(goi, sprintf("hsa — %s, %d dpi", ss$label, day), "hsa")
  }
}
rm(kegg_subsets)
```

# 10. Pathway visualisation (pathview)

```{r, message=FALSE, warning=FALSE}
# Add direction value for pathview colour mapping
for (day in 1:3) {
  obj_name <- paste0("combined_", day, "_jackknife_gallus")
  df <- get(obj_name)
  assign(obj_name, df %>% mutate(direction_value = ifelse(eff_direction == "+", 1, -1)))
}

pathway_ids <- c(
  'gga05164', # Influenza A
  'gga04620', # Toll-like receptor
  'gga04621', # NOD-like receptor
  'gga04622', # RIG-I-like receptor
  'gga04210', # Apoptosis
  'gga04217'  # Necroptosis
)

df_list <- list(
  day1 = combined_1_jackknife_gallus,
  day2 = combined_2_jackknife_gallus,
  day3 = combined_3_jackknife_gallus
)

for (day_label in names(df_list)) {
  df <- df_list[[day_label]]
  for (subset_type in c("jackknife", "combined")) {
    if (!subset_type %in% colnames(df)) next
    subset_df <- df[df[[subset_type]] == 1, c("entrezgene_id", "direction_value")]
    if (nrow(subset_df) == 0) next
    gene_data <- setNames(subset_df$direction_value, as.character(subset_df$entrezgene_id))
    for (pathway in pathway_ids) {
      tryCatch(
        pathview(gene.data = gene_data, pathway.id = pathway, species = "gga",
                 verbose = TRUE, kegg.native = TRUE,
                 out.suffix = paste0(day_label, "_", subset_type),
                 limit = list(gene = 2, cpd = 1),
                 low = list(gene = "red", cpd = "blue"),
                 mid = list(gene = "gray", cpd = "gray"),
                 high = list(gene = "green", cpd = "yellow")),
        error = function(e) {
          pathview(gene.data = gene_data, pathway.id = pathway, species = "gga",
                   verbose = TRUE, kegg.native = FALSE,
                   out.suffix = paste0(day_label, "_", subset_type, "_F"),
                   limit = list(gene = 2, cpd = 1),
                   low = list(gene = "red", cpd = "blue"),
                   mid = list(gene = "gray", cpd = "gray"),
                   high = list(gene = "green", cpd = "yellow"))
        }
      )
    }
  }
}
```

# 11. Comparison of KEGG pathways across studies (compareCluster)

```{r}
cluster_cols <- list(
  day1 = c("jackknife", "huang", "campbell", "smith"),
  day2 = c("jackknife", "huang", "campbell", "morris_2_3_2_1", "morris_2_2"),
  day3 = c("jackknife", "huang", "campbell", "smith")
)

long_title_base <- str_wrap(
  "Overrepresented KEGG Pathways In Differentially Expressed Genes Across Days Post Infection",
  width = 60
)

green_scale <- scale_fill_gradient(
  low  = rgb(188, 207, 0, maxColorValue = 255),
  high = rgb(58, 170, 53, maxColorValue = 255)
)

title_theme <- theme(plot.title = element_text(hjust = 0.5, size = 14, margin = margin(b = 10)))

run_kegg_compare <- function(df_list, cluster_cols, direction_filter = NULL, title_suffix = "") {
  for (day in names(df_list)) {
    df <- df_list[[day]]
    if (!is.null(direction_filter)) df <- df %>% filter(eff_direction == direction_filter)
    cols <- cluster_cols[[day]]
    geneclusters <- Filter(function(x) length(x) > 0, setNames(
      lapply(cols, function(col) {
        if (col %in% colnames(df)) as.character(df$entrezgene_id[df[[col]] == 1]) else NULL
      }), cols))
    if (length(geneclusters) == 0) { message(sprintf("No clusters for %s; skipping.", day)); next }
    cc <- compareCluster(geneclusters, fun = "enrichKEGG", organism = "gga", pvalueCutoff = 0.05)
    long_title <- str_wrap(paste0(long_title_base, title_suffix), width = 60)
    print(dotplot(cc) + ggtitle(long_title) + green_scale + title_theme)
  }
}

## 11.1 All genes (jackknife/study columns together)
run_kegg_compare(df_list, cluster_cols)

## 11.2 Upregulated only
run_kegg_compare(df_list, cluster_cols, direction_filter = "+",
                 title_suffix = " (Upregulated)")

## 11.3 Downregulated only
run_kegg_compare(df_list, cluster_cols, direction_filter = "-",
                 title_suffix = " (Downregulated)")
```

## 11.4 Jackknife genes across days (single compareCluster)

```{r}
geneclusters_days <- setNames(
  lapply(names(df_list), function(day) {
    df <- df_list[[day]]
    unique(as.character(df$entrezgene_id[df$jackknife == 1 & !is.na(df$entrezgene_id)]))
  }),
  c("Day 1", "Day 2", "Day 3")
)
geneclusters_days <- Filter(function(x) length(x) > 0, geneclusters_days)

cc_days <- compareCluster(geneclusters_days, fun = "enrichKEGG",
                          organism = "gga", pvalueCutoff = 0.05)
cc_days@compareClusterResult$Cluster <- factor(cc_days@compareClusterResult$Cluster,
                                               levels = names(geneclusters_days))
print(dotplot(cc_days) +
  ggtitle(str_wrap("KEGG Pathways in Jackknife-Positive Genes Across Days Post Infection", 60)) +
  green_scale + title_theme)

# Up/Down split across days
geneclusters6 <- unlist(lapply(seq_along(df_list), function(i) {
  df <- df_list[[i]]
  list(
    up   = unique(as.character(df$entrezgene_id[df$jackknife == 1 & df$eff_direction == '+'])),
    down = unique(as.character(df$entrezgene_id[df$jackknife == 1 & df$eff_direction == '-']))
  ) |> setNames(paste0("Day ", i, c(" (Up)", " (Down)")))
}), recursive = FALSE)
geneclusters6 <- Filter(function(x) length(x) > 0, geneclusters6)

cc6 <- compareCluster(geneclusters6, fun = "enrichKEGG", organism = "gga", pvalueCutoff = 0.05)
cc6@compareClusterResult$Cluster <- factor(cc6@compareClusterResult$Cluster, levels = names(geneclusters6))
print(dotplot(cc6) +
  ggtitle(str_wrap("KEGG Pathways in Jackknife-Positive Genes (Up/Down) Across Days", 60)) +
  green_scale + title_theme)
```

## 11.5 GO terms across days (custom GMT, enricher)

```{r}
# Build geneclusters for GO (Ensembl IDs from combined_*_jackknife)
df_list_go <- list(
  day1 = combined_1_jackknife,
  day2 = combined_2_jackknife,
  day3 = combined_3_jackknife
)

geneclusters_go <- setNames(
  lapply(df_list_go, function(df) {
    unique(as.character(na.omit(df$gene[df$jackknife == 1])))
  }),
  c("Day 1", "Day 2", "Day 3")
)
geneclusters_go <- Filter(function(x) length(x) > 0, geneclusters_go)

# Namespace-aware compareCluster helper
run_cc_namespace <- function(annotation_data, geneclusters, namespace_value,
                             pretty_ns = NULL, pvalueCutoff = 0.05,
                             qvalueCutoff = 0.20, minGSSize = 5) {
  stopifnot("namespace_1003" %in% colnames(annotation_data))
  if (is.null(pretty_ns)) pretty_ns <- tools::toTitleCase(gsub("_", " ", namespace_value))

  ann_ns <- annotation_data %>%
    dplyr::filter(!is.na(go_id), !is.na(ensembl_gene_id), namespace_1003 == namespace_value)
  if (nrow(ann_ns) == 0) { message("No rows for namespace '", namespace_value, "'"); return(invisible(NULL)) }

  TERM2GENE_ns <- ann_ns %>% dplyr::select(go_id, ensembl_gene_id) %>% distinct()
  TERM2NAME_ns <- if ("name_1006" %in% colnames(ann_ns))
    ann_ns %>% dplyr::select(go_id, name_1006) %>% distinct() else NULL

  bg <- unique(TERM2GENE_ns$ensembl_gene_id)

  cc_obj <- tryCatch(
    compareCluster(geneClusters = geneclusters, fun = "enricher",
                   TERM2GENE = TERM2GENE_ns, TERM2NAME = TERM2NAME_ns,
                   universe = bg, pAdjustMethod = "BH",
                   pvalueCutoff = pvalueCutoff, qvalueCutoff = qvalueCutoff,
                   minGSSize = minGSSize),
    error = function(e) { message("compareCluster error: ", e$message); NULL }
  )

  cc_df <- tryCatch(as.data.frame(cc_obj), error = function(e) NULL)
  if (is.null(cc_df) || nrow(cc_df) == 0) {
    message("No enrichment found for namespace '", namespace_value, "'")
    return(invisible(NULL))
  }

  cc_obj@compareClusterResult$Cluster <- factor(cc_obj@compareClusterResult$Cluster,
                                                 levels = names(geneclusters))
  p <- dotplot(cc_obj) +
    ggtitle(str_wrap(sprintf("GO %s — Jackknife Genes Across Days", pretty_ns), 60)) +
    green_scale + title_theme
  print(p)
  invisible(list(cc_object = cc_obj, cc_df = cc_df, plot = p))
}

ns_map <- c("biological_process" = "Biological Process",
            "molecular_function" = "Molecular Function",
            "cellular_component" = "Cellular Component")

res_bp <- run_cc_namespace(annotation_data, geneclusters_go, "biological_process", ns_map[1])
res_mf <- run_cc_namespace(annotation_data, geneclusters_go, "molecular_function", ns_map[2])
res_cc_go <- run_cc_namespace(annotation_data, geneclusters_go, "cellular_component", ns_map[3])
```

## 11.6 Common genes across all days

```{r}
genes_all_jackknife <- Reduce(intersect, lapply(1:3, function(d)
  get(paste0("combined_", d, "_jackknife_gallus")) %>%
    filter(jackknife == 1) %>% pull(gene)
))

build_day_df <- function(day, genes) {
  get(paste0("combined_", day, "_jackknife_gallus")) %>%
    filter(gene %in% genes) %>%
    dplyr::select(gene, gene_name, p_combined, FDR_combined, eff_direction) %>%
    rename_with(~ paste0(., "_day", day), c(p_combined, FDR_combined, eff_direction))
}

d1 <- build_day_df(1, genes_all_jackknife)
d2 <- build_day_df(2, genes_all_jackknife) %>% dplyr::select(-gene_name)
d3 <- build_day_df(3, genes_all_jackknife) %>% dplyr::select(-gene_name)

merged_jackknife <- d1 %>%
  inner_join(d2, by = "gene") %>%
  inner_join(d3, by = "gene") %>%
  relocate(gene_name, .after = gene) %>%
  filter(!is.na(gene_name) & gene_name != "")

datatable(merged_jackknife, rownames = FALSE)
rm(d1, d2, d3, build_day_df, genes_all_jackknife)
```

# 12. Publication figure — KEGG + GO panel

```{r final_enrichment_figure, warning=FALSE, message=FALSE, fig.width=15, fig.height=30}
suppressPackageStartupMessages({
  library(enrichplot)
  library(patchwork)
})

# KEGG: jackknife Entrez genes by day
geneclusters_kegg <- setNames(
  lapply(1:3, function(d) {
    df <- get(paste0("combined_", d, "_jackknife_gallus"))
    unique(na.omit(as.character(df$entrezgene_id[df$jackknife == 1])))
  }),
  c("Day_1", "Day_2", "Day_3")
)
geneclusters_kegg <- Filter(function(x) length(x) > 0, geneclusters_kegg)

cc_kegg <- compareCluster(geneclusters_kegg, fun = "enrichKEGG",
                           organism = "gga", pvalueCutoff = 0.05)
cc_kegg@compareClusterResult$Cluster <- factor(cc_kegg@compareClusterResult$Cluster,
                                               levels = names(geneclusters_kegg))

# GO BP: jackknife Ensembl genes by day
ann_bp <- annotation_data %>%
  filter(!is.na(go_id), !is.na(ensembl_gene_id), namespace_1003 == "biological_process")
TERM2GENE_BP <- ann_bp %>% dplyr::select(go_id, ensembl_gene_id) %>% distinct()
TERM2NAME_BP <- if ("name_1006" %in% colnames(ann_bp))
  ann_bp %>% dplyr::select(go_id, name_1006) %>% distinct() else NULL

geneclusters_gobp <- setNames(
  lapply(1:3, function(d) {
    df <- get(paste0("combined_", d, "_jackknife"))
    unique(na.omit(as.character(df$gene[df$jackknife == 1])))
  }),
  c("Day_1", "Day_2", "Day_3")
)
geneclusters_gobp <- Filter(function(x) length(x) > 0, geneclusters_gobp)

cc_go_bp <- compareCluster(
  geneclusters_gobp, fun = "enricher",
  TERM2GENE = TERM2GENE_BP, TERM2NAME = TERM2NAME_BP,
  universe = unique(TERM2GENE_BP$ensembl_gene_id),
  pAdjustMethod = "BH", pvalueCutoff = 0.05,
  qvalueCutoff = 0.20, minGSSize = 5
)
cc_go_bp@compareClusterResult$Cluster <- factor(cc_go_bp@compareClusterResult$Cluster,
                                                levels = names(geneclusters_gobp))

# Shared scale helpers
extract_minimal <- function(cc_obj) {
  df <- as.data.frame(cc_obj)
  df$GeneRatio_num <- sapply(df$GeneRatio, function(x) {
    parts <- strsplit(as.character(x), "/")[[1]]
    if (length(parts) != 2) return(NA_real_)
    as.numeric(parts[1]) / as.numeric(parts[2])
  })
  dplyr::select(df, p.adjust, GeneRatio = GeneRatio_num)
}

safe_range <- function(x, eps = 1e-6) {
  x <- x[is.finite(x)]
  if (length(x) == 0) return(c(0, 1))
  mn <- min(x); mx <- max(x)
  if (isTRUE(all.equal(mn, mx))) return(c(mn * (1 - eps), mx * (1 + eps)))
  c(mn, mx)
}

all_min   <- bind_rows(extract_minimal(cc_kegg), extract_minimal(cc_go_bp)) %>%
             filter(!is.na(p.adjust), !is.na(GeneRatio))
pval_range <- safe_range(all_min$p.adjust)
gr_range   <- safe_range(all_min$GeneRatio)

pub_fill <- scale_fill_gradient(
  low = rgb(188, 207, 0, maxColorValue = 255),
  high = rgb(58, 170, 53, maxColorValue = 255),
  limits = pval_range, name = "Adjusted p-value", na.value = "grey50"
)
pub_size <- scale_size(limits = gr_range, range = c(2, 8), name = "Gene Ratio")

p_kegg_pub <- dotplot(cc_kegg) +
  ggtitle("KEGG Pathways") + pub_fill + pub_size +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
        axis.text  = element_text(size = 9), legend.position = "right")

p_go_bp_pub <- dotplot(cc_go_bp) +
  ggtitle("GO Biological Processes") + pub_fill + pub_size +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
        axis.text  = element_text(size = 9), legend.position = "none")

final_pub_plot <- (p_kegg_pub | p_go_bp_pub) +
  plot_layout(guides = "collect", widths = c(1, 1)) +
  plot_annotation(
    title = str_wrap("Functional Enrichment of Jackknife-Positive Genes Across Days Post Infection", 80),
    tag_levels = "A",
    theme = theme(
      plot.title = element_text(hjust = 0.5, size = 16, face = "bold", margin = margin(b = 12)),
      plot.tag   = element_text(size = 14, face = "bold")
    )
  )

print(final_pub_plot)
ggsave("Figure_Functional_Enrichment_Jackknife.pdf", final_pub_plot,
       width = 12, height = 6, device = cairo_pdf)
ggsave("Figure_Functional_Enrichment_Jackknife.png", final_pub_plot,
       width = 12, height = 6, dpi = 600)

message("Figure exported: Figure_Functional_Enrichment_Jackknife.{pdf,png}")
```
